---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mixtureEM

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/pdvalencia/mixtureEM/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pdvalencia/mixtureEM/actions/workflows/R-CMD-check.yaml)

Latent class and latent profile analysis for applied researchers.

**mixtureEM** helps you find hidden subgroups in your data and describe
them (Collins & Lanza, 2010).

* If you have ever run an LCA, assigned everyone to their most likely
  class, and then run ANOVAs or chi-squares on the labels, this package
  exists for you. The step that is easiest to get wrong — relating classes
  to external variables — is the one mixtureEM automates correctly, using
  the bias-adjusted three-step estimators from the methodological
  literature (Bakk et al., 2014; Vermunt, 2010).
* If you have never run an LCA, this package exists for you too. Every
  step ships with a sensible default, and the output reads like the table
  you would put in a paper, so you do not need the estimation theory
  memorized to get correct answers out.

---

## Installation

You can install the development version from GitHub using the
[pak](https://pak.r-lib.org/) package:

```r
# install.packages("pak")
pak::pkg_install("pdvalencia/mixtureEM")
```

---

## Quick start

A complete analysis takes just a handful of calls. Using the bundled
`ventura_leon` data — 16 yes/no infidelity items from 400 young adults
(run `?ventura_leon` for dataset details) — you can fit a model and view
the class sizes:

```{r example}
library(mixtureEM)

items <- ventura_leon[, 7:22]   # the 16 infidelity items
fit <- fit_mixture(items, n_classes = 4, measurement = "binary",
                   n_init = 20, random_state = 1)
class_sizes(fit)
```

You can then easily plot the item-response probabilities to interpret the
classes:

```{r example-plot, fig.height = 5, fig.alt = "Item-response probabilities for the four fitted classes across the 16 infidelity items, one line per class."}
plot(fit, class_labels = c("Fidelity", "Affective interest",
                           "Infidelity", "Sexual desire"))
```

> Choosing how many classes to keep, reading the full item-response table,
> relating classes to covariates with a bias correction, and describing
> them by a distal outcome are each one function call away. The
> [`ventura_leon` worked
> example](https://pdvalencia.github.io/mixtureEM/articles/ventura_leon.html)
> walks through all of it on this exact fit.

---

## What it covers

| Analysis | Function | Worked example |
|---|---|---|
| Latent class analysis (binary / categorical / count items) | `fit_mixture()` | [`ventura_leon`](https://pdvalencia.github.io/mixtureEM/articles/ventura_leon.html) |
| Latent profile analysis (continuous items), missing data, multiple groups | `fit_mixture()` | [`janousch`](https://pdvalencia.github.io/mixtureEM/articles/janousch.html) |
| Choosing the number of classes (ICs, BLRT, fit diagnostics) | `compare_mixtures()`, `blrt()` | [`class_enumeration`](https://pdvalencia.github.io/mixtureEM/articles/class_enumeration.html) |
| Predictors of class membership (3-step ML) | `add_covariates()` | [`ventura_leon`](https://pdvalencia.github.io/mixtureEM/articles/ventura_leon.html) |
| Distal outcomes (BCH / ML) | `add_outcome()` | [`ventura_leon`](https://pdvalencia.github.io/mixtureEM/articles/ventura_leon.html) |
| Repeated-measures LCA (trajectory classes) | `fit_rmlca()` | [`rmlca`](https://pdvalencia.github.io/mixtureEM/articles/rmlca.html) |
| Latent transition analysis, mover-stayer models | `fit_lta()` | [`lta`](https://pdvalencia.github.io/mixtureEM/articles/lta.html) |
| Latent class growth analysis, growth mixture models | `fit_lcga()`, `fit_gmm()` | [`growth_mixture`](https://pdvalencia.github.io/mixtureEM/articles/growth_mixture.html) |
| Complex survey designs (weights, strata, clusters) | every model | [`survey_lca`](https://pdvalencia.github.io/mixtureEM/articles/survey_lca.html) |

Also included: mixed measurement models (binary + continuous + count
indicators in one model), full-information missing-data handling
everywhere, absolute and local fit diagnostics (`absolute_fit()`,
`bivariate_residuals()`), classification diagnostics, and standard errors
that account for the classes being estimated rather than observed
(`?covariate_se`).

---

## Design principles

* **Conceptual workflow.** Choose a model, *then* examine covariates and
  outcomes on that fitted model — `add_covariates(fit, ...)` and
  `add_outcome(fit, ...)` never re-estimate the classes you already
  inspected.
* **Sound defaults.** Stepwise analyses apply the recommended bias
  correction and uncertainty-aware standard errors unless you say
  otherwise; where a default has a literature, the default follows it.
* **Readable output, reusable numbers.** Summaries print as formatted
  tables *and* return tidy data frames.
* **No dependency stack.** Built using base R only, meaning it installs
  cleanly anywhere R does.

---

## Citation

If you use mixtureEM in published research, please cite it as:

> Valencia, P. D. (2026). mixtureEM: Latent class and profile analysis via
> mixture modelling (Version 0.2.0) [R package].
> <https://github.com/pdvalencia/mixtureEM>

## Acknowledgements

**mixtureEM** draws strong inspiration from the Python package
[**StepMix**](https://github.com/Labo-Lacourse/stepmix), which pioneered
open-source, bias-adjusted multi-step estimation of generalised mixture
models with external variables. If you make use of these methods, please
also consider citing the StepMix paper:

> Morin, S., Legault, R., Laliberte, F., Bakk, Z., Giguère, C.-E., de la
> Sablonnière, R., & Lacourse, E. (2025). StepMix: A Python package for
> pseudo-likelihood estimation of generalised mixture models with external
> variables. *Journal of Statistical Software*, *113*(8), 1–39.
> <https://doi.org/10.18637/jss.v113.i08>

## References

Bakk, Z., Oberski, D. L., & Vermunt, J. K. (2014). Relating latent class
assignments to external variables: Standard errors for correct inference.
*Political Analysis*, *22*(4), 520–540. <https://doi.org/10.1093/pan/mpu003>

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent transition
analysis: With applications in the social, behavioral, and health
sciences*. Wiley.

Vermunt, J. K. (2010). Latent class modeling with covariates: Two improved
three-step approaches. *Political Analysis*, *18*(4), 450–469.
<https://doi.org/10.1093/pan/mpq025>
