---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mixtureEM

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/pdvalencia/mixtureEM/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/pdvalencia/mixtureEM/actions/workflows/R-CMD-check.yaml)

**mixtureEM** is latent class and latent profile analysis for applied
researchers: find the hidden subgroups in your data, describe them, and
relate them to other variables — with methodologically sound defaults at
every step, in pure base R with no dependencies.

If you have ever run an LCA, assigned everyone to their most likely class,
and then run chi-squares on the labels, this package exists for you: the
step that is easiest to get wrong (relating classes to external variables)
is the one mixtureEM automates correctly, using the bias-adjusted
three-step estimators from the methodological literature (Vermunt, 2010;
Bakk, Oberski, & Vermunt, 2014).

## Installation

```r
# install.packages("remotes")
remotes::install_github("pdvalencia/mixtureEM")
```

## A complete analysis in a handful of calls

Using the bundled `ventura_leon` data — 16 yes/no infidelity items from
400 young adults (Ventura-León et al., 2025):

```{r example}
library(mixtureEM)

items <- ventura_leon[, 7:22]   # the 16 infidelity items

# 1. How many classes? Fit a range and compare.
selection <- compare_mixtures(items, k_range = 1:5, n_init = 10)

# 2. Fit the chosen model and look at it. BIC keeps creeping down past 4
#    classes, as it often does in LCA; we keep 4 on conceptual grounds,
#    following the original study's well-interpretable typology.
fit <- fit_mixture(items, n_classes = 4, n_init = 20, random_state = 1)
class_sizes(fit)
```

The item-class conditional probabilities — the table applied papers
report — come from `measurement_summary()`, which prints them and also
returns them as a data frame:

```{r example-params}
params <- measurement_summary(fit)
```

```{r example-plot, fig.height = 5}
plot(fit, class_labels = c("Fidelity", "Affective interest",
                           "Infidelity", "Sexual desire"))
```

```{r example-covariates}
# 3. Who ends up in which class? Bias-adjusted 3-step, on this exact fit.
fit_cov <- add_covariates(fit, ventura_leon[, c("sex", "age")])
summary(fit_cov)
```

(A reading note for the table above: a covariate can be significant on
its omnibus test while none of its printed rows are — the omnibus asks
whether the covariate distinguishes *any* pair of classes, while each row
only compares one class against the chosen reference class. Re-reference
with `summary(fit_cov, ref_class = 2)` to see the remaining contrasts.)

`add_outcome(fit, y)` asks the complementary question — how the classes
differ on an outcome.

## What it covers

| Analysis | Function | Worked example |
|---|---|---|
| Latent class analysis (binary / categorical / count items) | `fit_mixture()` | `vignette("ventura_leon")` |
| Latent profile analysis (continuous items), missing data, multiple groups | `fit_mixture()` | `vignette("janousch")` |
| Choosing the number of classes (ICs, BLRT, fit diagnostics) | `compare_mixtures()`, `blrt()` | `vignette("class_enumeration")` |
| Predictors of class membership (3-step ML) | `add_covariates()` | `vignette("ventura_leon")` |
| Distal outcomes (BCH / ML) | `add_outcome()` | `vignette("ventura_leon")` |
| Repeated-measures LCA (trajectory classes) | `fit_rmlca()` | `vignette("rmlca")` |
| Latent transition analysis, mover-stayer models | `fit_lta()` | `vignette("lta")` |
| Latent class growth analysis, growth mixture models | `fit_lcga()`, `fit_gmm()` | `vignette("growth_mixture")` |
| Complex survey designs (weights, strata, clusters) | every model | `vignette("survey_lca")` |

Also included: mixed measurement models (binary + continuous + count
indicators in one model), full-information missing-data handling
everywhere, absolute and local fit diagnostics (`absolute_fit()`,
`bivariate_residuals()`), classification diagnostics, and standard errors
that account for the classes being estimated rather than observed
(`?covariate_se`).

## Design principles

* **Conceptual workflow.** Choose a model, *then* examine covariates and
  outcomes on that fitted model — `add_covariates(fit, ...)` and
  `add_outcome(fit, ...)` never re-estimate the classes you already
  inspected.
* **Sound defaults.** Stepwise analyses apply the recommended bias
  correction and uncertainty-aware standard errors unless you say
  otherwise; where a default has a literature, the default follows it.
* **Readable output, reusable numbers.** Summaries print as formatted
  tables *and* return tidy data frames.
* **No dependency stack.** Base R only, so it installs anywhere R does.

## Citation

If you use mixtureEM in published research, please cite it as:

```
Valencia, P. D. (2026). mixtureEM: Latent class and profile analysis via
mixture modelling (Version 0.2.0) [R package].
https://github.com/pdvalencia/mixtureEM
```

## Acknowledgements

**mixtureEM** draws strong inspiration from the Python package
[**StepMix**](https://github.com/Labo-Lacourse/stepmix), which pioneered
open-source, bias-adjusted multi-step estimation of generalised mixture
models with external variables. If you make use of these methods, please
also consider citing the StepMix paper:

> Morin, S., Legault, R., Laliberte, F., Bakk, Z., Giguère, C.-E., de la
> Sablonnière, R., & Lacourse, E. (2025). StepMix: A Python package for
> pseudo-likelihood estimation of generalised mixture models with external
> variables. *Journal of Statistical Software*, *113*(8), 1–39.
> <https://doi.org/10.18637/jss.v113.i08>

## References

Bakk, Z., Oberski, D. L., & Vermunt, J. K. (2014). Relating latent class
assignments to external variables: Standard errors for correct inference.
*Political Analysis*, *22*(4), 520–540. <https://doi.org/10.1093/pan/mpu003>

Vermunt, J. K. (2010). Latent class modeling with covariates: Two improved
three-step approaches. *Political Analysis*, *18*(4), 450–469.
<https://doi.org/10.1093/pan/mpq025>
