---
title: "Choosing the Number of Classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the Number of Classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
has_polca <- requireNamespace("poLCA", quietly = TRUE)
```

```{r setup}
library(mixtureEM)
```

Every latent class analysis starts with the same question: *how many
classes?* There is no single decisive statistic, so the standard practice
is to fit a range of solutions and weigh several kinds of evidence
together (Nylund et al., 2007):

1. **Information criteria** (BIC, SABIC, AIC) — lower is better.
2. **The bootstrap likelihood ratio test (BLRT)** — a formal test of
   *k* classes against *k* − 1.
3. **Classification quality** (entropy, average posterior probabilities).
4. **Absolute and local fit** — does the chosen model actually reproduce
   the data?
5. **Interpretability and class sizes** — a smallest class under 5% of
   the sample, or with fewer than about 25 cases, is one you have to be
   able to defend on substantive grounds, however much it improves the
   BIC (Lee et al., 2023). That is a reporting convention, not a rule;
   mixtureEM applies no minimum and will happily estimate a class of
   three people.

This vignette walks through all five on a classic dataset:
`poLCA::carcinoma`, in which seven pathologists each rated the same 118
biopsy slides for the presence of carcinoma (Agresti, 2002). The question
a latent class model asks here is: how many *types of slide* are there,
given how the raters agree and disagree?

The chunks below run only if the `poLCA` package is installed
(`install.packages("poLCA")`).

## The data

```{r data, eval = has_polca}
data(carcinoma, package = "poLCA")
# poLCA codes ratings 1/2; recode to 0/1 for measurement = "binary".
ratings <- as.matrix(carcinoma) - 1
head(ratings)
```

## Information criteria

`compare_mixtures()` fits each candidate number of classes and prints the
information criteria side by side:

```{r ics, eval = has_polca}
set.seed(1)
selection <- compare_mixtures(ratings, k_range = 1:5,
                              measurement = "binary", n_init = 20)
```

BIC prefers three classes here — the solution the methodological
literature has repeatedly recovered for these data: clearly negative
slides, clearly positive slides, and an ambiguous group the raters split
on.

## The bootstrap likelihood ratio test

The BLRT asks the sharper question "does adding one more class fit
significantly better?", with the null distribution generated by
parametric bootstrap (McLachlan & Peel, 2000). It is slow, because each
replication refits both models; keep `n_reps` modest while exploring. A
test you intend to *report* is a different matter: refit it at
`n_reps = 999`. A bootstrap p-value can only take the values
1/(B + 1), 2/(B + 1), …, so at 50 or 100 draws it cannot separate .04
from .06, and `blrt()` warns when the result lands that close to the
line. See `vignette("estimation")` for the power figures behind those
numbers.

```{r blrt, eval = has_polca, cache = TRUE}
set.seed(2)
blrt_23 <- blrt(ratings, k_small = 2, k_large = 3,
                measurement = "binary", n_reps = 50, verbose = FALSE)
blrt_23

set.seed(3)
blrt_34 <- blrt(ratings, k_small = 3, k_large = 4,
                measurement = "binary", n_reps = 50, verbose = FALSE)
blrt_34
```

The 2-vs-3 test rejects (a third class is needed); the 3-vs-4 test does
not, agreeing with BIC.

## Classification quality and fit of the chosen model

```{r chosen, eval = has_polca}
set.seed(4)
fit <- fit_mixture(ratings, n_classes = 3, n_init = 30)
fit
class_sizes(fit)
classification_diagnostics(fit)
```

### Which entropy is being reported

Relative entropy is an absolute entropy expressed on a 0-1 scale, and there is
more than one convention for the denominator that puts it there. mixtureEM
divides by $n \log K$, the entropy of a uniform distribution over the $K$
classes — the largest value the absolute entropy could take. Some other
implementations divide instead by $n H(\hat\pi)$, the entropy of the *estimated*
class sizes. Both appear in the literature and neither is wrong, but they do not
give the same number, so an entropy quoted from elsewhere is not always
comparable with the one printed here.

Because $H(\hat\pi) \le \log K$ always, with equality only when the classes come
out exactly equal in size, the figure reported here is always the higher of the
two. The gap is small for balanced solutions and grows as the class sizes become
uneven. To convert:

$$R^2_{\text{other}} = 1 - (1 - R^2_{\text{here}}) \frac{\log K}{H(\hat\pi)},
\qquad H(\pi) = -\sum_k \pi_k \log \pi_k$$

Either way, entropy describes how cleanly a solution separates cases; it is not
evidence for how many classes there are. A model can separate two classes
beautifully and still be the wrong model.

Absolute fit compares the observed response-pattern frequencies with the
model-implied ones, and the bivariate residuals localise any misfit to
specific pairs of items:

```{r fit-checks, eval = has_polca}
absolute_fit(fit)
bivariate_residuals(fit)
```

Finally, the profile plot is the interpretability check — three classes
that are well separated *and* substantively meaningful:

```{r plot, eval = has_polca}
plot(fit, class_labels = c("Negative", "Positive", "Ambiguous"),
     main = "Pathologist agreement (3-class LCA)")
```

## References

Agresti, A. (2002). *Categorical data analysis* (2nd ed.). Wiley.

Lee, T. K., Wickrama, K. A. S., & O'Neal, C. W. (2023). An introduction
to growth mixture models (GMM). In *International Encyclopedia of
Education* (4th ed., Vol. 14, pp. 646–655). Elsevier.
<https://doi.org/10.1016/B978-0-12-818630-5.10076-4>

McLachlan, G., & Peel, D. (2000). *Finite mixture models*. Wiley.

Sclove, S. L. (1987). Application of model-selection criteria to some
problems in multivariate analysis. *Psychometrika*, *52*(3), 333–343.
<https://doi.org/10.1007/BF02294360>

Nylund, K. L., Asparouhov, T., & Muthén, B. O. (2007). Deciding on the
number of classes in latent class analysis and growth mixture modeling: A
Monte Carlo simulation study. *Structural Equation Modeling*, *14*(4),
535–569. <https://doi.org/10.1080/10705510701575396>
