---
title: "What mixtureEM Estimates, and Why"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{What mixtureEM Estimates, and Why}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mixtureEM)
```

Most of the time you do not need this vignette. The defaults are the settings
the package intends you to use, and the other vignettes show the analyses.

This one exists for the moments when a number does not look the way you
expected: a log-likelihood that differs from another program's, a warning about
a collapsed variance, a probability reported as 0.000. Each of those is a
deliberate choice with a literature behind it, and each is easier to live with
once you know which choice it was.

## Mixture likelihoods have more than one maximum

Fitting a mixture model means climbing a surface with several peaks. The EM
algorithm climbs the one it starts nearest, so a single run is a coin toss
rather than an estimate. `fit_mixture()` therefore runs 20 random starts by
default and keeps the best, and it reports how many of them found that best
solution:

```{r starts}
set.seed(1)
n <- 400
p <- matrix(c(.9, .85, .8, .15,
              .1, .20, .15, .90), nrow = 2, byrow = TRUE)
z <- sample(1:2, n, TRUE)
X <- matrix(rbinom(n * 4, 1, p[z, ]), n, 4)
colnames(X) <- paste0("q", 1:4)

fit <- fit_mixture(X, n_classes = 2, measurement = "binary", random_state = 1)
fit
```

Read that line as a stability check. If the best solution was found by most of
the starts, the surface has been mapped and more starts will not change the
answer. If it was found **once**, raise `n_init` and refit: a maximum seen a
single time may just be the best of a small sample of the surface, and a
different seed could beat it. This is the standard multi-start report in the
latent class literature (Hipp & Bauer, 2006; Nylund-Gibson & Choi, 2018).

Set `random_state` to make the search reproducible.

## The estimator is a posterior mode, not maximum likelihood

Maximum likelihood in latent class analysis routinely pushes a conditional
probability to exactly 0 or 1. When that happens the information matrix loses
rank, and standard errors, confidence intervals and significance tests for those
parameters stop meaning anything (Galindo Garre & Vermunt, 2006, sec. 2.1).

mixtureEM avoids that with weak priors that hold estimates strictly inside the
parameter space — the *Bayes constants* approach, controlled by four named
counts:

```{r bayes-constants, eval = FALSE}
bayes_constants = list(latent = 1, categorical = 1, poisson = 1, variances = 1)
```

Each is a count of artificial observations spread over the classes, so its
influence shrinks as the sample grows and is negligible on any class with real
data behind it. All four default to 1.

Three consequences worth knowing:

- **The reported log-likelihood is the log-likelihood**, evaluated at the
  posterior mode — not the log-posterior. It is the same function a
  maximum-likelihood program reports, evaluated at slightly different estimates,
  so it is bounded above by the ML value. The gap grows with the number of
  parameters ML would have placed on the boundary, and it keeps AIC and BIC on a
  defensible scale.
- **Setting a constant to `0` recovers plain maximum likelihood** for that block.
  That is an escape hatch for reproducing an unregularized reference analysis,
  not a recommended setting.
- **`measurement_summary()` flags probabilities that reached the boundary
  anyway.** Those tell you something real — a class defined partly by an item
  every case in it answers the same way — and warn you not to read their
  standard errors.

## Continuous indicators: the likelihood is unbounded

For continuous indicators the problem is sharper. Put a class mean on a few
identical cases and shrink its variance towards zero, and the density at those
cases grows without limit: the likelihood is *unbounded*, and the highest-scoring
solution can be one that describes nothing (Day, 1969; Kiefer & Wolfowitz, 1956).

When a fit lands there, mixtureEM says so and names the item and the class. The
important thing about that warning is what it is **not**: it is not a convergence
failure. Raising `n_init` will not fix it and can make it worse, because a longer
search finds a taller spike. For the same reason a flagged fit's BIC cannot be
compared with a clean one's — it is inflated by the spike.

There are three ways out, and they change the model in different ways:

- `variances_equal = TRUE` holds each item's variance equal across classes. The
  likelihood is then bounded and the problem cannot arise. This is the
  homoscedastic latent-profile model and the constrained approach of Hathaway
  (1985).
- **Fewer classes.** A class describing a spike rather than a subgroup usually
  means the data do not support this many.
- **A stronger prior**, `bayes_constants = list(variances = ...)`. A useful
  starting point is one artificial observation per class — that is
  `variances = n_classes` — increasing it a little at a time if the warning
  persists. The warning prints that number for your model.

Then check the result rather than just that the warning stopped: the flagged
variance should no longer be far below the others, and its class mean should
have come off the floor or ceiling of the response scale.

Two things about that prior are worth knowing, because they are what make the
advice a rule rather than a number to copy.

It is centred on each item's *own* observed marginal variance, not on a fixed
quantity. That is what lets one constant mean the same thing on a five-point
scale and on an income variable: rescale an indicator and the prior rescales
with it. A penalty pinned to a constant has no such property, and the
penalised-likelihood literature prefers the data-scaled form for exactly this
reason (Chen, Tan & Zhang, 2008, sec. 4).

And the constant is divided among the classes, so `variances = n_classes` is
the setting that holds it at one artificial observation per class at *every*
number of classes — the way the literature applies it, the same amount to every
component. A bare number does the opposite: it is a different amount of prior in
a three-class model than in a six-class one. Write it as `n_classes` and it
stays comparable across the models you are enumerating.

Theory settles the *form* of this prior and not its size. Within the conditions
Chen, Tan & Zhang (2008) impose — which any fixed positive constant meets — a
penalty that diverges as a variance approaches zero, and grows more slowly than
the sample, gives a consistent estimator whatever its constant. That result is
proved for *univariate* normal mixtures; the multivariate case, which is what
this package fits, they leave open (sec. 5). So the choice of constant is all
the more a finite-sample judgement, which is why the default is deliberately
weak and the advice above comes with a check attached.

### A clean fit is not proof that there is no spike

The absence of the warning does not establish that no collapsed solution exists.
The warning can only report on the solution the search actually reached, and
some degenerate solutions sit in basins so narrow that random starts never land
in them. A fit can replicate its maximum on every one of hundreds of starts, and
still not be the highest point on the likelihood.

This is worth checking whenever an indicator has a floor or a ceiling — a
bounded scale with a visible pile-up of cases at an endpoint. That pile-up is a
set of near-identical values, which is exactly what a collapsing class latches
onto, but it attracts only starts that begin very close to it. Two cheap checks:
refit with `bayes_constants = list(variances = n_classes)` and see whether the
solution moves at all, and look at whether any class mean is sitting on a scale
endpoint with a variance far below that item's marginal. Splitting the sample
into subgroups widens these basins, because the pile-up makes up a larger share
of each smaller sample, so a model that looks clean on the whole sample deserves
the check again on the parts.

## Missing data needs nothing from you

Any indicator containing `NA` is estimated by full-information maximum
likelihood under the usual missing-at-random assumption, selected
automatically. Cases missing on *every* indicator are dropped, with a warning
saying how many, because they contribute nothing while still counting towards
the sample size in BIC.

## Multiple groups: what the log-likelihood is over

With `group =`, mixtureEM maximises the likelihood of the indicators **given**
the group. The grouping variable's own distribution is not modelled and its
proportions are not counted as parameters.

That is a convention, and programs differ. Software that treats a fully observed
grouping variable as a latent class variable instead adds the group's
multinomial term to the log-likelihood and its `G - 1` proportions to the
parameter count. Both differences are fixed constants: they cancel in any
likelihood-ratio test and in any comparison among models fitted here. For
reading numbers off such output directly, a `group` fit also carries
`metrics$ll_knownclass` and `metrics$n_params_knownclass`.

## References

Chen, J., Tan, X., & Zhang, R. (2008). Inference for normal mixtures in mean and
variance. *Statistica Sinica*, *18*(2), 443-465.

Day, N. E. (1969). Estimating the components of a mixture of normal
distributions. *Biometrika*, *56*(3), 463-474.

Galindo Garre, F., & Vermunt, J. K. (2006). Avoiding boundary estimates in
latent class analysis by Bayesian posterior mode estimation. *Behaviormetrika*,
*33*(1), 43-59.

Hathaway, R. J. (1985). A constrained formulation of maximum-likelihood
estimation for normal mixture distributions. *The Annals of Statistics*, *13*(2),
795-800.

Hipp, J. R., & Bauer, D. J. (2006). Local solutions in the estimation of growth
mixture models. *Psychological Methods*, *11*(1), 36-53.

Kiefer, J., & Wolfowitz, J. (1956). Consistency of the maximum likelihood
estimator in the presence of infinitely many incidental parameters. *The Annals
of Mathematical Statistics*, *27*(4), 887-906.

McLachlan, G. J., & Peel, D. (2000). *Finite Mixture Models*. Wiley.

Nylund-Gibson, K., & Choi, A. Y. (2018). Ten frequently asked questions about
latent class analysis. *Translational Issues in Psychological Science*, *4*(4),
440-461.
