---
title: "Trajectory Classes: LCGA and Growth Mixture Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trajectory Classes: LCGA and Growth Mixture Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

When a continuous outcome is measured repeatedly, two related models group
cases by the *shape* of their trajectory over time, rather than by a
single score:

* **Latent class growth analysis (LCGA)** sorts cases into classes that
  each follow one shared curve — everyone in a class is assumed to follow
  essentially the same trajectory, give or take measurement noise (Jung &
  Wickrama, 2008).
* **Growth mixture models (GMM)** relax that assumption: cases still
  belong to a class with its own typical curve, but each person is
  allowed to start a bit higher or lower, and rise or fall a bit faster or
  slower, than their class's average (Muthén & Shedden, 1999). LCGA is the
  special case where that person-to-person wiggle room is fixed at zero.

The practical difference: because LCGA has no other way to absorb
within-class differences, it tends to need *more*, narrower classes to
fit the same data; a GMM can often describe it with fewer, broader ones.

## Simulated example: five-wave symptom scores

We simulate 700 cases in two known trajectory classes — a stable-low class
and a declining class — with genuine within-class variation in both
intercept and slope, i.e. GMM-type data:

```{r simulate}
set.seed(2026)
n     <- 700
cls   <- rbinom(n, 1, 0.4)                       # 1 = Declining (40%)
icept <- ifelse(cls == 1, 6.0, 2.0) + rnorm(n, 0, 0.9)
slope <- ifelse(cls == 1, -0.7, 0.05) + rnorm(n, 0, 0.25)
waves <- 0:4
Y <- outer(icept, rep(1, 5)) + outer(slope, waves) +
  matrix(rnorm(n * 5, 0, 0.8), n, 5)
colnames(Y) <- paste0("score_t", 1:5)
```

## LCGA first

An LCGA with a linear curve per class:

```{r lcga}
fit_lcga2 <- fit_lcga(Y, times = 5, n_classes = 2, family = "gaussian",
                      degree = 1, n_init = 15, random_state = 5)
fit_lcga2
```

## The GMM

`fit_gmm()` adds random intercepts and slopes (the default,
`random_effects = "intercept_slope"`):

```{r gmm, cache = TRUE}
fit_gmm2 <- fit_gmm(Y, times = 5, n_classes = 2, n_init = 15,
                    random_state = 5)
fit_gmm2
```

Because LCGA is nested in the GMM, the information criteria say directly
whether the random effects are earning their parameters — here they are,
by a wide margin, as simulated:

```{r compare}
data.frame(
  model = c("LCGA (2 classes)", "GMM (2 classes)"),
  ll    = c(fit_lcga2$metrics$ll, fit_gmm2$metrics$ll),
  bic   = c(fit_lcga2$metrics$bic, fit_gmm2$metrics$bic)
)
```

`plot()` overlays the observed class means on the estimated curves — the
check that a class describes a real subgroup rather than an artifact of
splitting one cloud in half:

```{r plot}
plot(fit_gmm2, class_labels = c("Stable low", "Declining"))
```

## Covariates, two different questions

Two kinds of covariate enter a growth mixture model, and they answer
different questions:

* **Who is in which class?** — class-membership predictors, through the
  usual stepwise workflow:

```{r membership}
x_member <- rnorm(n, mean = 0.7 * cls)
fit_cov  <- add_covariates(fit_gmm2, x_member)
results  <- summary(fit_cov)
```

* **Who, within a class, starts higher or declines faster?** — covariates
  on the growth factors themselves, which are part of the measurement
  model and therefore supplied to `fit_gmm()` directly:

```{r growth-predictors, cache = TRUE}
x_growth <- rnorm(n)
fit_gp   <- fit_gmm(Y, times = 5, n_classes = 2,
                    growth_predictors = x_growth,
                    n_init = 15, random_state = 5)
print(fit_gp)
```

## Practical cautions

* A GMM whose data cannot support its covariance structure does not fail
  loudly — it drives a variance to the boundary. mixtureEM warns when a
  solution sits on the boundary; treat that as a sign to simplify
  (`psi = "equal"`, `random_effects = "intercept"`, or LCGA).
* Class enumeration is harder than in cross-sectional LCA: BIC, the
  plotted trajectories against observed means, and substantive
  interpretability should all agree before a solution is reported
  (Nylund et al., 2007).

## References

Jung, T., & Wickrama, K. A. S. (2008). An introduction to latent class
growth analysis and growth mixture modeling. *Social and Personality
Psychology Compass*, *2*(1), 302–317.
<https://doi.org/10.1111/j.1751-9004.2007.00054.x>

Muthén, B., & Shedden, K. (1999). Finite mixture modeling with mixture
outcomes using the EM algorithm. *Biometrics*, *55*(2), 463–469.
<https://doi.org/10.1111/j.0006-341X.1999.00463.x>

Nylund, K. L., Asparouhov, T., & Muthén, B. O. (2007). Deciding on the
number of classes in latent class analysis and growth mixture modeling: A
Monte Carlo simulation study. *Structural Equation Modeling*, *14*(4),
535–569. <https://doi.org/10.1080/10705510701575396>
