---
title: "Trajectory Classes: LCGA and Growth Mixture Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trajectory Classes: LCGA and Growth Mixture Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

When a continuous outcome is measured repeatedly, two related models group
cases by the *shape* of their trajectory over time, rather than by a
single score:

* **Latent class growth analysis (LCGA)** sorts cases into classes that
  each follow one shared curve — everyone in a class is assumed to follow
  essentially the same trajectory, give or take measurement noise (Jung &
  Wickrama, 2008).
* **Growth mixture models (GMM)** relax that assumption: cases still
  belong to a class with its own typical curve, but each person is
  allowed to start a bit higher or lower, and rise or fall a bit faster or
  slower, than their class's average (Muthén & Shedden, 1999). LCGA is the
  special case where that person-to-person wiggle room is fixed at zero.

The practical difference: because LCGA has no other way to absorb
within-class differences, it tends to need *more*, narrower classes to
fit the same data; a GMM can often describe it with fewer, broader ones.

## Simulated example: five-wave symptom scores

We simulate 700 cases in two known trajectory classes — a stable-low class
and a declining class — with genuine within-class variation in both
intercept and slope, i.e. GMM-type data:

```{r simulate}
set.seed(2026)
n     <- 700
cls   <- rbinom(n, 1, 0.4)                       # 1 = Declining (40%)
icept <- ifelse(cls == 1, 6.0, 2.0) + rnorm(n, 0, 0.9)
slope <- ifelse(cls == 1, -0.7, 0.05) + rnorm(n, 0, 0.25)
waves <- 0:4
Y <- outer(icept, rep(1, 5)) + outer(slope, waves) +
  matrix(rnorm(n * 5, 0, 0.8), n, 5)
colnames(Y) <- paste0("score_t", 1:5)
```

## LCGA first

An LCGA with a linear curve per class:

```{r lcga}
fit_lcga2 <- fit_lcga(Y, times = 5, n_classes = 2, family = "gaussian",
                      degree = 1, n_init = 15, random_state = 5)
fit_lcga2
```

## The GMM

`fit_gmm()` adds random intercepts and slopes (the default,
`random_effects = "intercept_slope"`):

```{r gmm, cache = TRUE}
fit_gmm2 <- fit_gmm(Y, times = 5, n_classes = 2, n_init = 15,
                    random_state = 5)
fit_gmm2
```

Because LCGA is nested in the GMM, the information criteria say directly
whether the random effects are earning their parameters — here they are,
by a wide margin, as simulated:

```{r compare}
data.frame(
  model = c("LCGA (2 classes)", "GMM (2 classes)"),
  ll    = c(fit_lcga2$metrics$ll, fit_gmm2$metrics$ll),
  bic   = c(fit_lcga2$metrics$bic, fit_gmm2$metrics$bic)
)
```

`plot()` overlays the observed class means on the estimated curves — the
check that a class describes a real subgroup rather than an artifact of
splitting one cloud in half:

```{r plot}
plot(fit_gmm2, class_labels = c("Stable low", "Declining"))
```

## Covariates, two different questions

Two kinds of covariate enter a growth mixture model, and they answer
different questions:

* **Who is in which class?** — class-membership predictors, through the
  usual stepwise workflow:

```{r membership}
x_member <- rnorm(n, mean = 0.7 * cls)
fit_cov  <- add_covariates(fit_gmm2, x_member)
results  <- summary(fit_cov)
```

* **Who, within a class, starts higher or declines faster?** — covariates
  on the growth factors themselves, which are part of the measurement
  model and therefore supplied to `fit_gmm()` directly:

```{r growth-predictors, cache = TRUE}
x_growth <- rnorm(n)
fit_gp   <- fit_gmm(Y, times = 5, n_classes = 2,
                    growth_predictors = x_growth,
                    n_init = 15, random_state = 5)
print(fit_gp)
```

## Reporting the model

`measurement_summary()` returns the growth parameters as a tidy data
frame — the growth-factor means, their variances and covariances, the
residual variances and the fitted trajectory, one row per parameter and
class — so the numbers can go into a table without being retyped from the
printed output:

```{r summary}
params <- measurement_summary(fit_gmm2)
head(params[params$parameter == "growth_mean", ])
```

Class enumeration has its own two functions. `compare_longitudinal()`
fits a range of class counts with one specification and tabulates the
criteria, and for a growth model its default range starts at one class —
the ordinary latent growth curve model, which is the benchmark the class
solutions have to beat:

```{r enumerate, eval = FALSE}
compare_longitudinal(Y, model = "gmm", times = 5, k_range = 1:3,
                     n_init = 15, random_state = 5)
```

`blrt()` takes the fitted model itself, so the bootstrapped comparison is
run on the same time design and covariance structure rather than on a
re-specified model:

```{r blrt, eval = FALSE}
blrt(k_small = 2, k_large = 3, from_fit = fit_gmm2, n_reps = 100)
```

## Practical cautions

* A GMM whose data cannot support its covariance structure does not fail
  loudly — it drives a variance towards zero. mixtureEM warns when a
  residual variance or a class's growth-factor variation has collapsed
  relative to the observed variance of the outcome; treat that as a sign
  to simplify (`psi = "equal"`, `random_effects = "intercept"`, fewer
  classes, or LCGA). Two things about that warning are worth taking
  literally. The collapsed solution is usually the one with the *best*
  BIC, because a variance heading for zero inflates the likelihood
  without bound, so BIC cannot be used to choose between it and a clean
  fit. And it is not a convergence failure, so raising `n_init` will not
  fix it: more starts means more chances of finding the spike, and a
  solution that only one start in fifty reaches is evidence about the
  specification rather than about the search (Hipp & Bauer, 2006).
* **A lower log-likelihood is not always a worse answer.** The
  growth-factor covariance is constrained here to be positive
  semi-definite, and the residual variances to be positive. Software that
  does not impose those constraints can report a negative variance, or
  variances that are each positive while implying a correlation greater
  than one in absolute value — and the likelihood it prints for such a
  solution is evaluated outside the parameter space, so it is not
  comparable with one that is inside it. When mixtureEM reports a lower
  likelihood than another program on the same model, check the other
  program's covariance matrix before concluding anything: an admissible
  solution with a smaller likelihood is the reportable one. The right
  response to a model that cannot produce a proper solution is Berlin,
  Parra and Williams's (2014): to report that no reasonable and proper
  solution was found, and to simplify the specification.
* Class enumeration is harder than in cross-sectional LCA: BIC, the
  plotted trajectories against observed means, and substantive
  interpretability should all agree before a solution is reported
  (Nylund et al., 2007). Checklists for reporting a trajectory study
  (van de Schoot et al., 2017) ask for the one-class solution, the
  criteria at every class count, the class sizes and the classification
  diagnostics; `compare_longitudinal()`, `class_sizes()` and
  `classification_diagnostics()` supply them.

## References

Berlin, K. S., Parra, G. R., & Williams, N. A. (2014). An introduction to
latent variable mixture modeling (part 2): Longitudinal latent class
growth analysis and growth mixture models. *Journal of Pediatric
Psychology*, *39*(2), 188–203. <https://doi.org/10.1093/jpepsy/jst085>

Hipp, J. R., & Bauer, D. J. (2006). Local solutions in the estimation of
growth mixture models. *Psychological Methods*, *11*(1), 36–53.
<https://doi.org/10.1037/1082-989X.11.1.36>

Jung, T., & Wickrama, K. A. S. (2008). An introduction to latent class
growth analysis and growth mixture modeling. *Social and Personality
Psychology Compass*, *2*(1), 302–317.
<https://doi.org/10.1111/j.1751-9004.2007.00054.x>

Lee, T. K., Wickrama, K. A. S., & O'Neal, C. W. (2023). An introduction to
growth mixture models (GMM). In *International Encyclopedia of Education*
(4th ed., Vol. 14, pp. 646–655). Elsevier.
<https://doi.org/10.1016/B978-0-12-818630-5.10076-4>

Muthén, B., & Shedden, K. (1999). Finite mixture modeling with mixture
outcomes using the EM algorithm. *Biometrics*, *55*(2), 463–469.
<https://doi.org/10.1111/j.0006-341X.1999.00463.x>

Ram, N., & Grimm, K. J. (2009). Growth mixture modeling: A method for
identifying differences in longitudinal change among unobserved groups.
*International Journal of Behavioral Development*, *33*(6), 565–576.
<https://doi.org/10.1177/0165025409343765>

Nylund, K. L., Asparouhov, T., & Muthén, B. O. (2007). Deciding on the
number of classes in latent class analysis and growth mixture modeling: A
Monte Carlo simulation study. *Structural Equation Modeling*, *14*(4),
535–569. <https://doi.org/10.1080/10705510701575396>

van de Schoot, R., Sijbrandij, M., Winter, S. D., Depaoli, S., & Vermunt,
J. K. (2017). The GRoLTS-checklist: Guidelines for reporting on latent
trajectory studies. *Structural Equation Modeling*, *24*(3), 451–467.
<https://doi.org/10.1080/10705511.2016.1247646>
