---
title: "Resilience Profiles Across Countries: Multi-Group LPA and Measurement Invariance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resilience Profiles Across Countries: Multi-Group LPA and Measurement Invariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## Background

This vignette walks through Janousch, Anyan, Kassis, Morote, Hjemdal,
Sidler, Graf, Rietz, Chouvati, & Govaris (2022), *"Resilience Profiles
Across Context: A Latent Profile Analysis in a German, Greek, and Swiss
Sample of Adolescents,"* PLOS ONE, 17(1):e0263089
(<https://doi.org/10.1371/journal.pone.0263089>), using the `janousch`
dataset bundled with mixtureEM (`?janousch`; CC BY 4.0, figshare).

The study measured 1,160 seventh-graders from Germany, Greece, and
Switzerland on two symptom indicators (anxiety, depression) and five
protective-factor subscales from the Resilience Scale for Adolescents
(personal competence, social competence, structured style, social
resources, family cohesion). It then asked two questions we replicate
here: (1) how many resilience profiles best describe each country's data,
and (2) are those profiles *measurement-invariant* across countries — that
is, do the same profile labels mean the same thing (the same underlying
symptom/protective-factor levels) in each country?

The original analysis handled the substantial item-level missingness with
Mplus's full-information ML estimator (MLR), rather than deleting
incomplete cases; we do the same here using `fit_mixture()`'s
`"continuous_nan"` measurement type, which models missing continuous
indicators directly rather than requiring complete cases.

## The data

```{r data}
data(janousch)
str(janousch)
```

```{r items}
items <- c("anxiety", "depression", "personal_competence", "social_competence",
           "structured_style", "social_resources", "family_cohesion")
```

## Per-country profiles

The paper fit a separate LPA for each country and selected, by BIC and
interpretability, 3 profiles for Switzerland and 4 for Germany and Greece.
Finding the best solution in a model with missing continuous indicators
benefits from many random restarts (the paper itself used up to 2,000);
we use a more modest `n_init` here for speed, but recommend increasing it
for a real analysis.

```{r fit-switzerland}
set.seed(1)
fit_ch <- fit_mixture(janousch[janousch$country == "Switzerland", items],
                       n_classes = 3, measurement = "continuous_nan",
                       n_init = 30, max_iter = 2000)
class_sizes(fit_ch)
```

This closely recovers the paper's Swiss solution (Non-Resilient 22.1%,
Moderately Resilient 42.9%, Untroubled 34.9%). What the profiles *mean*
comes from the class-specific indicator means, which
`measurement_summary()` prints and also returns as a data frame:

```{r profile-ch}
profile_ch <- measurement_summary(fit_ch)
```

```{r fit-germany}
set.seed(1)
fit_de <- fit_mixture(janousch[janousch$country == "Germany", items],
                       n_classes = 4, measurement = "continuous_nan",
                       n_init = 30, max_iter = 2000)
class_sizes(fit_de)
```

This closely recovers the paper's German solution (Non-Resilient 15.7%,
Moderately Resilient 44.2%, Untroubled 27.3%, Resilient 12.7%).

```{r fit-greece}
set.seed(1)
fit_gr <- fit_mixture(janousch[janousch$country == "Greece", items],
                       n_classes = 4, measurement = "continuous_nan",
                       n_init = 30, max_iter = 2000)
class_sizes(fit_gr)
```

For Greece, our solution's class sizes differ more noticeably from the
paper's (Non-Resilient 21.0%, Moderately Resilient 30.8%, Untroubled 24.9%,
Resilient 23.3%). The Greek sample had the lowest entropy in the original
study (.788, vs. .81-.84 for the other two countries), meaning its classes
are less well separated — exactly the situation where an LPA's likelihood
surface has multiple close local optima and different software (or
different random starts) can land on distinct, similarly-plausible
solutions. This is a useful reminder that LPA/LCA solutions, especially
lower-entropy ones, should be treated as one reasonable answer rather than
*the* answer, and checked for stability across many random starts.

## Testing measurement invariance across countries

Rather than three separate models, mixtureEM can fit all three countries
*jointly* as a multiple-group model. Fitting the same number of classes
(4, following the paper's own supplementary invariance analysis, which
tested the four-profile solution across all three countries) with
`group_effects = "both"` gives each country's own fully separate
item-response profile (a "configural" model); `group_effects =
"measurement"` instead holds the item-response means equal across
countries, only letting class *sizes* differ. Comparing the two with a
likelihood-ratio test is a direct test of measurement invariance.

```{r invariance, cache = TRUE}
set.seed(1)
fit_configural <- fit_mixture(janousch[items], n_classes = 4,
                               measurement = "continuous_nan",
                               group = janousch$country,
                               group_effects = "both",
                               n_steps = 1, n_init = 50, max_iter = 2000)

set.seed(1)
fit_invariant <- fit_mixture(janousch[items], n_classes = 4,
                              measurement = "continuous_nan",
                              group = janousch$country,
                              group_effects = "measurement",
                              n_steps = 1, n_init = 50, max_iter = 2000)

longitudinal_lrt(fit_invariant, fit_configural)
```

The restriction is rejected: forcing the same profile means across
countries fits significantly worse than letting each country have its own.
This matches the paper's own conclusion ("Measurement invariance did not
hold across the three countries") — the same four profile *labels*
(Non-Resilient, Moderately Resilient, Untroubled, Resilient) do not
correspond to the same underlying levels of symptoms and protective
factors in each country, so cross-country comparisons of these profiles
should be made cautiously.

## Covariates: gender and migration background

The paper related class membership to gender and migration background
using Mplus's R3STEP procedure — itself already a 3-step, classification-
uncertainty-aware approach (equivalent to mixtureEM's `correction = "ML"`),
so unlike the Ventura-León vignette, there is no weaker baseline to
improve on here. We reproduce the same kind of analysis for Switzerland,
where the paper reported no significant covariate effects, as a check:

The Swiss model was already fit above, so `add_covariates()` runs the
three-step analysis on that exact solution — no re-specification and no
re-estimation of the measurement model:

```{r covariates}
ch <- janousch[janousch$country == "Switzerland", ]
fit_ch_cov <- add_covariates(fit_ch, ch[, c("gender", "migration_background")])
summary(fit_ch_cov)
```

One detail worth noticing: the full dataset codes `gender` with three
levels, but no Swiss adolescent chose "Other", so after subsetting that
level is empty. mixtureEM drops empty levels automatically (with a
message) rather than printing a degenerate coefficient row for a category
with no data — and it warns when an *observed* level rests on only a
handful of cases, whose odds ratio would be unstable.

## References

Janousch, C., Anyan, F., Kassis, W., Morote, R., Hjemdal, O., Sidler, P.,
Graf, U., Rietz, C., Chouvati, R., & Govaris, C. (2022). Resilience
profiles across context: A latent profile analysis in a German, Greek, and
Swiss sample of adolescents. *PLOS ONE*, *17*(1), e0263089.
<https://doi.org/10.1371/journal.pone.0263089>
