---
title: "Resilience Profiles Across Countries: Multi-Group LPA and Measurement Invariance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resilience Profiles Across Countries: Multi-Group LPA and Measurement Invariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## Background

This vignette walks through Janousch et al. (2022), *"Resilience Profiles
Across Context: A Latent Profile Analysis in a German, Greek, and Swiss
Sample of Adolescents,"* PLOS ONE, 17(1):e0263089
(<https://doi.org/10.1371/journal.pone.0263089>), using the `janousch`
dataset bundled with mixtureEM (`?janousch`; CC BY 4.0, figshare).

The study measured 1,160 seventh-graders from Germany, Greece, and
Switzerland on two symptom indicators (anxiety, depression) and five
protective-factor subscales from the Resilience Scale for Adolescents
(personal competence, social competence, structured style, social
resources, family cohesion). It then asked two questions we replicate
here: (1) how many resilience profiles best describe each country's data,
and (2) are those profiles *measurement-invariant* across countries — that
is, do the same profile labels mean the same thing (the same underlying
symptom/protective-factor levels) in each country?

The original analysis handled the substantial item-level missingness with a
full-information ML estimator rather than deleting incomplete cases, and so do
we — without having to ask for it. `fit_mixture()` detects the missing values
and models them directly; `measurement = "continuous"` is all that is needed.
Every fit below also passes `variances_equal = FALSE`. A continuous measurement
model now holds each item's variance equal across the classes by default,
because that bounds the likelihood; this analysis deliberately frees them,
since the variance collapse that follows is the thing the vignette is about.
In practical terms: instead of dropping an adolescent who skipped one item —
which is what listwise deletion, the default in many packages, would do — the
model uses every answer that adolescent *did* give.

## The data

```{r data}
data(janousch)
str(janousch)
```

```{r items}
items <- c("anxiety", "depression", "personal_competence", "social_competence",
           "structured_style", "social_resources", "family_cohesion")
```

## Per-country profiles

The paper fit a separate LPA for each country and selected, by BIC and
interpretability, 3 profiles for Switzerland and 4 for Germany and Greece.
Finding the best solution in a model with missing continuous indicators
benefits from many random restarts (the paper itself used up to 2,000);
we use a more modest `n_init` here for speed, but recommend increasing it
for a real analysis.

```{r fit-switzerland}
set.seed(1)
fit_ch <- fit_mixture(janousch[janousch$country == "Switzerland", items],
                       n_classes = 3, measurement = "continuous", variances_equal = FALSE,
                       n_init = 30, max_iter = 2000)
class_sizes(fit_ch)
```

This closely recovers the paper's Swiss solution (Non-Resilient 22.1%,
Moderately Resilient 42.9%, Untroubled 34.9%). What the profiles *mean*
comes from the class-specific indicator means, which
`measurement_summary()` prints and also returns as a data frame:

```{r profile-ch}
profile_ch <- measurement_summary(fit_ch)
```

Profiles are easier to name than to read off a table, so plot them. `plot()`
puts every indicator on a common 0–1 axis (continuous items are min–max scaled
against their observed range, which is what the `*` marks), so the *shape* of
each line — high on the protective factors, low on the symptoms, or the
reverse — is what you interpret:

```{r plot-ch}
plot(fit_ch, main = "Resilience profiles, Switzerland")
```

Once you have decided what each line represents, read the profile means off
`profile_ch` to fix the class order, then pass your own names in that order
through `class_labels = c(...)`; the legend will carry them, with each class's
size attached. Those names — "Non-Resilient", "Moderately Resilient",
"Untroubled" — are the paper's, and this plot is how they were arrived at.

```{r fit-germany}
set.seed(1)
fit_de <- fit_mixture(janousch[janousch$country == "Germany", items],
                       n_classes = 4, measurement = "continuous", variances_equal = FALSE,
                       n_init = 30, max_iter = 2000)
class_sizes(fit_de)
```

This closely recovers the paper's German solution (Non-Resilient 15.7%,
Moderately Resilient 44.2%, Untroubled 27.3%, Resilient 12.7%).

```{r fit-greece}
set.seed(1)
fit_gr0 <- fit_mixture(janousch[janousch$country == "Greece", items],
                       n_classes = 4, measurement = "continuous", variances_equal = FALSE,
                       n_init = 30, max_iter = 2000)
```

That fit raises a warning: one class's variance on `social_resources` has
collapsed towards zero. The next section explains what that means and why it
happens; the short version is that a mixture of normals with freely estimated
variances has an *unbounded* likelihood, so a class can score well by sitting on
a handful of identical answers instead of describing a subgroup. In survey terms
this is usually a **ceiling or floor effect**: when a large block of respondents
picks the very top of the `social_resources` scale, the model can manufacture a
"profile" out of those identical scores alone. Such a solution is not
interpretable, and — importantly — more random starts will not rescue it.

> **If your variance collapses**
>
> 1. **Read the warning.** It names the item and the class (here,
>    `social_resources` in one of the four Greek classes) and suggests a
>    starting value.
> 2. **Look at that item.** A histogram usually shows the culprit: a heavy pile
>    of responses on the highest or lowest scale point.
> 3. **Raise the prior, not `n_init`.** Start at the suggested
>    `bayes_constants = list(variances = ...)` and increase it a little at a
>    time until the warning goes away.
> 4. **Re-run every model you intend to compare at the same value**, so the
>    fits stay on a common footing.

The remedy is that stronger prior holding the variances away from zero. It is
best read as a gentle statement of disbelief that any real subgroup has exactly
zero spread: `variances = 5` adds the equivalent of five hypothetical
observations, spread across the classes, whose only job is to smooth the
mathematical spike away. The warning suggests a starting point of one artificial
observation per class — `variances = 4` for this four-class model — and says to
raise it if the warning persists, which on this data it does. A little more is
enough:

```{r fit-greece-clean}
set.seed(1)
fit_gr <- fit_mixture(janousch[janousch$country == "Greece", items],
                       n_classes = 4, measurement = "continuous", variances_equal = FALSE,
                       n_init = 30, max_iter = 2000,
                       bayes_constants = list(variances = 5))
class_sizes(fit_gr)
```

Even with a clean fit, the Greek class sizes differ more noticeably from the
paper's (Non-Resilient 21.0%, Moderately Resilient 30.8%, Untroubled 24.9%,
Resilient 23.3%) than the Swiss and German ones did. The Greek sample had the
lowest entropy in the original study (.788, vs. .81-.84 for the other two
countries), meaning its classes are less well separated — exactly the situation
where an LPA's likelihood surface has several close local optima and different
software can land on distinct, similarly plausible solutions. Two different
things are at work here and they call for different responses: *local optima*
are a search problem, answered by more random starts, and the printed "best
solution found by k of n starts" line tells you whether you have enough; a
*collapsed variance* is not a search problem at all, and more starts make it
worse.

## Testing measurement invariance across countries

Before we can say "Greece has more Resilient adolescents than Germany", we have
to establish that *being* Resilient means the same thing in both places. If the
profile means differ by country, the labels are attached to different underlying
patterns and the comparison is apples to oranges. That is the question
measurement invariance answers, and it is worth settling before any
cross-country claim goes into a manuscript.

Rather than three separate models, mixtureEM can fit all three countries
*jointly* as a multiple-group model, and ask directly whether a profile
means the same thing in each. Fitting the same number of classes (4,
following the paper's own supplementary invariance analysis, which tested
the four-profile solution across all three countries), the comparison is
between two nested models:

| Model | Profile means | Indicator variances | Class sizes | Arguments |
|---|---|---|---|---|
| Invariant | pooled | pooled | free by country | `group_effects = "prevalence"` |
| Non-invariant | **free by country** | pooled | free by country | `group_effects = "both"`, `group_invariant_params = "covariances"` |

Measurement invariance is the restriction that a profile sits at the same
place on every indicator in every country. Freeing the *means* by country
is what breaks it; the indicator variances stay pooled in both models,
because they describe how tightly people cluster around a profile rather
than where the profile is, and holding them equal is what keeps the
comparison a test of the profiles themselves. This is the standard
latent-profile invariance comparison (Olivera-Aguilar & Rikoon, 2018), and
it is also the smaller and better-behaved of the models available: a fully
heterogeneous alternative, which frees the variances by country as well, is
fitted at the end of this section for comparison.

Let us fit the restricted model first, exactly as one would by default:

```{r invariance-default, cache = TRUE}
fit_invariant <- fit_mixture(janousch[items], n_classes = 4,
                             measurement = "continuous", variances_equal = FALSE,
                             group = janousch$country,
                             group_effects = "prevalence",
                             n_steps = 1, n_init = 50, max_iter = 2000,
                             random_state = 11)
```

That fit raises the same warning the Greek model did: one class's variance on
`social_resources` has **collapsed** towards zero, with the class mean sitting
exactly at the top of the scale.

This is worth understanding rather than working around, because it is a
property of the model and not a bug in the data. A mixture of normals with
freely estimated variances has an *unbounded* likelihood: take any class,
shrink its variance on one item towards zero, and put its mean on a few
cases that happen to share a value, and the density at those cases — and
so the likelihood — grows without limit. Such a solution can score better
than any meaningful one while describing nothing but a clump of identical
answers. Here, enough adolescents report the maximum on `social_resources`
that a class can form on the ceiling alone.

Two practical consequences follow, and both cut against instinct. Raising
`n_init` does **not** help: the likelihood really is unbounded in that
direction, so a longer search finds a taller spike and reports a better
log-likelihood for a worse solution. And a flagged fit's BIC cannot be compared
with a clean one's, because it is inflated by that spike.

mixtureEM guards against this with a weak prior that holds variances away from
zero. The default is deliberately mild, and as the Greek fit showed, mild enough
that it does not prevent every collapse — its job is to keep the problem finite
and *visible*, so the warning can name the item and the class, which it does.
The remedy is the same stronger prior, at the same strength the Greek fit needed:

```{r invariance, cache = TRUE}
fit_invariant <- fit_mixture(janousch[items], n_classes = 4,
                             measurement = "continuous", variances_equal = FALSE,
                             group = janousch$country,
                             group_effects = "prevalence",
                             n_steps = 1, n_init = 50, max_iter = 2000,
                             random_state = 11,
                             bayes_constants = list(variances = 5))

fit_free_means <- fit_mixture(janousch[items], n_classes = 4,
                              measurement = "continuous", variances_equal = FALSE,
                              group = janousch$country,
                              group_effects = "both",
                              group_invariant_params = "covariances",
                              n_steps = 1, n_init = 50, max_iter = 2000,
                              random_state = 11,
                              bayes_constants = list(variances = 5))

lr_test(fit_invariant, fit_free_means)
```

Both fits are now clean — no collapsed variances in either — and the
restriction is decisively rejected: forcing the same profile means across
countries fits significantly worse than letting each country have its own.
This matches the paper's own conclusion ("Measurement invariance did not
hold across the three countries") — the same four profile *labels*
(Non-Resilient, Moderately Resilient, Untroubled, Resilient) do not
correspond to the same underlying levels of symptoms and protective
factors in each country.

For the write-up, that has a concrete consequence: report and interpret the
profiles **separately within each country**, and do not pool the three samples
or compare profile prevalences across them as if the labels were interchangeable.
A statement like "the Untroubled profile is more common in Greece than in
Germany" is not supported by these data, because "Untroubled" is not the same
profile in the two countries.

One caution about the comparison itself: `bayes_constants` was raised on
*both* models, which is what makes them comparable. A likelihood-ratio
test between two fits estimated under different priors is not a test of
anything.

### If you also want the variances to differ

`group_invariant_params = "covariances"` is what holds the indicator
variances pooled while the means move. Dropping it frees everything about
the measurement model, giving each country its own profile means *and* its
own variances:

```{r configural, cache = TRUE}
fit_configural <- fit_mixture(janousch[items], n_classes = 4,
                              measurement = "continuous", variances_equal = FALSE,
                              group = janousch$country,
                              group_effects = "both",
                              n_steps = 1, n_init = 50, max_iter = 2000,
                              random_state = 11,
                              bayes_constants = list(variances = 5))

lr_test(fit_invariant, fit_configural)
```

This rejects too, and by more, but it is answering a broader question —
"is anything about the measurement model different across countries?"
rather than "do the profiles sit in the same place?" — on twice the
degrees of freedom. Report it when heterogeneity of *spread* is itself of
interest; otherwise the means-only test above is the one that matches the
substantive claim.

It is also the harder model to fit: 177 parameters spread across three
country-specific blocks, which random starting values struggle to bring
into correspondence. mixtureEM warm-starts it from the per-country
solutions to help, but a larger `n_init` may still improve it, which would
make the statistic larger and the rejection stronger, not weaker.
`lr_test()` warns if the search has gone badly enough to make the
full model score *below* the restricted one.

## Covariates: gender and migration background

The paper related class membership to gender and migration background
using Mplus's R3STEP procedure — itself already a 3-step, classification-
uncertainty-aware approach (equivalent to mixtureEM's `correction = "ML"`).
We reproduce the same kind of analysis for Switzerland,
where the paper reported no significant covariate effects, as a check:

The Swiss model was already fit above, so `add_covariates()` runs the
three-step analysis on that exact solution — no re-specification and no
re-estimation of the measurement model:

```{r covariates}
ch <- janousch[janousch$country == "Switzerland", ]
fit_ch_cov <- add_covariates(fit_ch, ch[, c("gender", "migration_background")])
summary(fit_ch_cov)
```

Each row is one predictor level in one contrast: the class named in the header
versus the reference class (Class 1). The odds ratio says how the odds of
falling in that class rather than the reference change for that level of the
predictor, holding the other covariates fixed. A `gender.Female` row with an
odds ratio near 1 and a p-value far above .05 therefore reads as: girls are no
more or less likely than boys to land in that profile rather than the reference
one. Values above 1 favour the named class, values below 1 favour the reference
class, and only the p-value tells you whether the difference is distinguishable
from none — an odds ratio of 1.5 on a p of .6 is not evidence of anything.

One detail worth noticing: the full dataset codes `gender` with three
levels, but no Swiss adolescent chose "Other", so after subsetting that
level is empty. mixtureEM drops empty levels automatically (with a
message) rather than printing a degenerate coefficient row for a category
with no data — and it warns when an *observed* level rests on only a
handful of cases, whose odds ratio would be unstable.

## References

Janousch, C., Anyan, F., Kassis, W., Morote, R., Hjemdal, O., Sidler, P.,
Graf, U., Rietz, C., Chouvati, R., & Govaris, C. (2022). Resilience
profiles across context: A latent profile analysis in a German, Greek, and
Swiss sample of adolescents. *PLOS ONE*, *17*(1), e0263089.
<https://doi.org/10.1371/journal.pone.0263089>

Olivera-Aguilar, M., & Rikoon, S. H. (2018). Assessing measurement
invariance in multiple-group latent profile analysis. *Structural Equation
Modeling*, *25*(3), 439–452.
<https://doi.org/10.1080/10705511.2017.1408015>
