---
title: "Resilience Profiles Across Countries: Multi-Group LPA and Measurement Invariance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resilience Profiles Across Countries: Multi-Group LPA and Measurement Invariance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## Background

This vignette walks through Janousch et al. (2022), *"Resilience Profiles
Across Context: A Latent Profile Analysis in a German, Greek, and Swiss
Sample of Adolescents,"* PLOS ONE, 17(1):e0263089
(<https://doi.org/10.1371/journal.pone.0263089>), using the `janousch`
dataset bundled with mixtureEM (`?janousch`; CC BY 4.0, figshare).

The study measured 1,160 seventh-graders from Germany, Greece, and
Switzerland on two symptom indicators (anxiety, depression) and five
protective-factor subscales from the Resilience Scale for Adolescents
(personal competence, social competence, structured style, social
resources, family cohesion). It then asked two questions we replicate
here: (1) how many resilience profiles best describe each country's data,
and (2) are those profiles *measurement-invariant* across countries — that
is, do the same profile labels mean the same thing (the same underlying
symptom/protective-factor levels) in each country?

The original analysis handled the substantial item-level missingness with
Mplus's full-information ML estimator (MLR), rather than deleting
incomplete cases; we do the same here using `fit_mixture()`'s
`"continuous_nan"` measurement type, which models missing continuous
indicators directly rather than requiring complete cases.

## The data

```{r data}
data(janousch)
str(janousch)
```

```{r items}
items <- c("anxiety", "depression", "personal_competence", "social_competence",
           "structured_style", "social_resources", "family_cohesion")
```

## Per-country profiles

The paper fit a separate LPA for each country and selected, by BIC and
interpretability, 3 profiles for Switzerland and 4 for Germany and Greece.
Finding the best solution in a model with missing continuous indicators
benefits from many random restarts (the paper itself used up to 2,000);
we use a more modest `n_init` here for speed, but recommend increasing it
for a real analysis.

```{r fit-switzerland}
set.seed(1)
fit_ch <- fit_mixture(janousch[janousch$country == "Switzerland", items],
                       n_classes = 3, measurement = "continuous_nan",
                       n_init = 30, max_iter = 2000)
class_sizes(fit_ch)
```

This closely recovers the paper's Swiss solution (Non-Resilient 22.1%,
Moderately Resilient 42.9%, Untroubled 34.9%). What the profiles *mean*
comes from the class-specific indicator means, which
`measurement_summary()` prints and also returns as a data frame:

```{r profile-ch}
profile_ch <- measurement_summary(fit_ch)
```

```{r fit-germany}
set.seed(1)
fit_de <- fit_mixture(janousch[janousch$country == "Germany", items],
                       n_classes = 4, measurement = "continuous_nan",
                       n_init = 30, max_iter = 2000)
class_sizes(fit_de)
```

This closely recovers the paper's German solution (Non-Resilient 15.7%,
Moderately Resilient 44.2%, Untroubled 27.3%, Resilient 12.7%).

```{r fit-greece}
set.seed(1)
fit_gr <- fit_mixture(janousch[janousch$country == "Greece", items],
                       n_classes = 4, measurement = "continuous_nan",
                       n_init = 30, max_iter = 2000)
class_sizes(fit_gr)
```

For Greece, our solution's class sizes differ more noticeably from the
paper's (Non-Resilient 21.0%, Moderately Resilient 30.8%, Untroubled 24.9%,
Resilient 23.3%). The Greek sample had the lowest entropy in the original
study (.788, vs. .81-.84 for the other two countries), meaning its classes
are less well separated — exactly the situation where an LPA's likelihood
surface has multiple close local optima and different software (or
different random starts) can land on distinct, similarly-plausible
solutions. This is a useful reminder that LPA/LCA solutions, especially
lower-entropy ones, should be treated as one reasonable answer rather than
*the* answer, and checked for stability across many random starts.

## Testing measurement invariance across countries

Rather than three separate models, mixtureEM can fit all three countries
*jointly* as a multiple-group model, and ask directly whether a profile
means the same thing in each. Fitting the same number of classes (4,
following the paper's own supplementary invariance analysis, which tested
the four-profile solution across all three countries), the comparison is
between two nested models:

| Model | Profile means | Indicator variances | Class sizes | Arguments |
|---|---|---|---|---|
| Invariant | pooled | pooled | free by country | `group_effects = "prevalence"` |
| Non-invariant | **free by country** | pooled | free by country | `group_effects = "both"`, `group_invariant_params = "covariances"` |

Measurement invariance is the restriction that a profile sits at the same
place on every indicator in every country. Freeing the *means* by country
is what breaks it; the indicator variances stay pooled in both models,
because they describe how tightly people cluster around a profile rather
than where the profile is, and holding them equal is what keeps the
comparison a test of the profiles themselves. This is the standard
latent-profile invariance comparison (Olivera-Aguilar & Rikoon, 2018), and
it is also the smaller and better-behaved of the models available: a fully
heterogeneous alternative, which frees the variances by country as well, is
fitted at the end of this section for comparison.

Let us fit the restricted model first, exactly as one would by default:

```{r invariance-default, cache = TRUE}
fit_invariant <- fit_mixture(janousch[items], n_classes = 4,
                             measurement = "continuous_nan",
                             group = janousch$country,
                             group_effects = "prevalence",
                             n_steps = 1, n_init = 50, max_iter = 2000,
                             random_state = 11)
```

That fit raises a warning: one class's variance on `social_resources` has
**collapsed** towards zero, with the class mean sitting exactly at the top
of the scale.

This is worth understanding rather than working around, because it is a
property of the model and not a bug in the data. A mixture of normals with
freely estimated variances has an *unbounded* likelihood: take any class,
shrink its variance on one item towards zero, and put its mean on a few
cases that happen to share a value, and the density at those cases — and
so the likelihood — grows without limit. Such a solution can score better
than any meaningful one while describing nothing but a clump of identical
answers. Here, enough adolescents report the maximum on `social_resources`
that a class can form on the ceiling alone.

mixtureEM guards against this with a weak prior that holds variances away
from zero, and the default strength is deliberately mild — mild enough to
leave the per-country solutions above exactly where the paper found them.
Its job is not to prevent every collapse but to keep it finite and
*visible*, so the warning can name the item and the class, which it does.
The remedy it recommends is a stronger prior:

```{r invariance, cache = TRUE}
fit_invariant <- fit_mixture(janousch[items], n_classes = 4,
                             measurement = "continuous_nan",
                             group = janousch$country,
                             group_effects = "prevalence",
                             n_steps = 1, n_init = 50, max_iter = 2000,
                             random_state = 11,
                             bayes_constants = list(variances = 5))

fit_free_means <- fit_mixture(janousch[items], n_classes = 4,
                              measurement = "continuous_nan",
                              group = janousch$country,
                              group_effects = "both",
                              group_invariant_params = "covariances",
                              n_steps = 1, n_init = 50, max_iter = 2000,
                              random_state = 11,
                              bayes_constants = list(variances = 5))

longitudinal_lrt(fit_invariant, fit_free_means)
```

Both fits are now clean — no collapsed variances in either — and the
restriction is decisively rejected: forcing the same profile means across
countries fits significantly worse than letting each country have its own.
This matches the paper's own conclusion ("Measurement invariance did not
hold across the three countries") — the same four profile *labels*
(Non-Resilient, Moderately Resilient, Untroubled, Resilient) do not
correspond to the same underlying levels of symptoms and protective
factors in each country, so cross-country comparisons of these profiles
should be made cautiously.

One caution about the comparison itself: `bayes_constants` was raised on
*both* models, which is what makes them comparable. A likelihood-ratio
test between two fits estimated under different priors is not a test of
anything.

### If you also want the variances to differ

`group_invariant_params = "covariances"` is what holds the indicator
variances pooled while the means move. Dropping it frees everything about
the measurement model, giving each country its own profile means *and* its
own variances:

```{r configural, cache = TRUE}
fit_configural <- fit_mixture(janousch[items], n_classes = 4,
                              measurement = "continuous_nan",
                              group = janousch$country,
                              group_effects = "both",
                              n_steps = 1, n_init = 50, max_iter = 2000,
                              random_state = 11,
                              bayes_constants = list(variances = 5))

longitudinal_lrt(fit_invariant, fit_configural)
```

This rejects too, and by more, but it is answering a broader question —
"is anything about the measurement model different across countries?"
rather than "do the profiles sit in the same place?" — on twice the
degrees of freedom. Report it when heterogeneity of *spread* is itself of
interest; otherwise the means-only test above is the one that matches the
substantive claim.

It is also the harder model to fit: 177 parameters spread across three
country-specific blocks, which random starting values struggle to bring
into correspondence. mixtureEM warm-starts it from the per-country
solutions to help, but a larger `n_init` may still improve it, which would
make the statistic larger and the rejection stronger, not weaker.
`longitudinal_lrt()` warns if the search has gone badly enough to make the
full model score *below* the restricted one.

## Covariates: gender and migration background

The paper related class membership to gender and migration background
using Mplus's R3STEP procedure — itself already a 3-step, classification-
uncertainty-aware approach (equivalent to mixtureEM's `correction = "ML"`).
We reproduce the same kind of analysis for Switzerland,
where the paper reported no significant covariate effects, as a check:

The Swiss model was already fit above, so `add_covariates()` runs the
three-step analysis on that exact solution — no re-specification and no
re-estimation of the measurement model:

```{r covariates}
ch <- janousch[janousch$country == "Switzerland", ]
fit_ch_cov <- add_covariates(fit_ch, ch[, c("gender", "migration_background")])
summary(fit_ch_cov)
```

One detail worth noticing: the full dataset codes `gender` with three
levels, but no Swiss adolescent chose "Other", so after subsetting that
level is empty. mixtureEM drops empty levels automatically (with a
message) rather than printing a degenerate coefficient row for a category
with no data — and it warns when an *observed* level rests on only a
handful of cases, whose odds ratio would be unstable.

## References

Janousch, C., Anyan, F., Kassis, W., Morote, R., Hjemdal, O., Sidler, P.,
Graf, U., Rietz, C., Chouvati, R., & Govaris, C. (2022). Resilience
profiles across context: A latent profile analysis in a German, Greek, and
Swiss sample of adolescents. *PLOS ONE*, *17*(1), e0263089.
<https://doi.org/10.1371/journal.pone.0263089>

Olivera-Aguilar, M., & Rikoon, S. H. (2018). Assessing measurement
invariance in multiple-group latent profile analysis. *Structural Equation
Modeling*, *25*(3), 439–452.
<https://doi.org/10.1080/10705511.2017.1408015>
