---
title: "Latent Profile Analysis: Bystander Responses to Sexual Harassment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent Profile Analysis: Bystander Responses to Sexual Harassment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## What this dataset is

`liang_ark_sim` is simulated data. It is drawn from the published parameters
of Liang and Ark's Study 2 three-class latent profile solution, not from
actual survey responses, and it contains no real bystander accounts. Fit
statistics, standard errors and p-values computed on it below are properties
of this simulated dataset, not of Liang and Park's study, and should not be
cited as findings about workplace bystander behavior. The simulation
reproduces the published solution closely where the indicators are concerned
— refitted profile means land within about 0.15 of a scale point of the
published ones, and entropy comes out near the published .905 — and only
approximately for the covariates, for the reason given in the covariate
section below.

Two columns exist only because the data are synthetic and no real dataset
would carry them: `class_true`, the profile each case was actually drawn
from, and `boundary`, which flags cases drawn from a blend of two profiles
rather than one. This vignette deliberately does not look at either column
until the very end, fitting the model exactly as an applied researcher
would on real data.

```{r data}
data(liang_ark_sim)
str(liang_ark_sim)
```

The five latent profile indicators are bystander actions in response to
witnessing sexual harassment at work: `confront` (direct confrontation of
the harasser), `distract` (distraction or interruption), `support`
(emotional support to the target), `report` (reporting to an authority),
and `discuss` (speaking with the target afterwards).

## How many profiles?

Equal (class-invariant) variances first — this is what the original study
fitted and what `fit_mixture()` defaults to.

```{r enumerate, cache = TRUE}
ind <- liang_ark_sim[, c("confront", "distract", "support", "report", "discuss")]
set.seed(1)
comp <- compare_mixtures(ind, k_range = 1:5, measurement = "continuous",
                         variances_equal = TRUE, n_init = 20)
plot(comp)
```

BIC keeps improving out to five classes here, which is common with
information criteria on a modest sample: the criterion never turns a
corner and instead trades off against a solution getting harder to
interpret and to replicate. The published analysis specified three
profiles, and the section below carries that forward.

## What happens when you free the variances

```{r free-variance, cache = TRUE}
set.seed(2)
free_v <- fit_mixture(ind, n_classes = 3, measurement = "continuous",
                      variances_equal = FALSE, n_init = 30)
```

Freeing the residual variances lets a class shrink onto a handful of cases
and drive its own variance toward zero, which sends the likelihood to
infinity and makes the solution uninterpretable; the package warns exactly
this happened above. The prior on the variances (default strength 1) stops
this from happening when set stronger:

```{r free-variance-fixed, cache = TRUE}
set.seed(2)
free_v_fixed <- fit_mixture(ind, n_classes = 3, measurement = "continuous",
                            variances_equal = FALSE, n_init = 30,
                            bayes_constants = list(variances = 3))
free_v_fixed
```

Setting the prior's strength to the number of classes is the setting that
behaves the same way whatever K is. Even so, the equal-variance model is
the one this vignette carries forward: it is what the published analysis
specified, and parsimony is a real argument here, not a consolation prize
for the free model being illegitimate.

## The three-profile solution

```{r fit, cache = TRUE}
fit <- fit_mixture(ind, n_classes = 3, measurement = "continuous",
                   variances_equal = TRUE, n_init = 20, random_state = 7)
fit
params <- measurement_summary(fit)
```

Classes come out sorted by size. Reading `params`: class 1 is high on
`support` and unremarkable elsewhere, class 2 is low on all five
indicators, and class 3 is high on all five. That is *supportive only*,
*disengaged*, and *broad responders*, in that order:

```{r plot, fig.height = 5.5}
plot(fit, type = "bar",
     class_labels = c("Supportive only", "Disengaged", "Broad responders"),
     main = "Bystander response profiles")
```

These three labels are this vignette's, chosen to describe the response
pattern. Liang and Park name the same three profiles after the level of
personal risk each one involves: what is called *Supportive only* here is their
**low-risk intervention** profile, *Broad responders* is their **active
intervention** profile, and *Disengaged* is their **no/limited intervention**
profile. The profiles are the same; only the names differ. Use theirs when
reporting against the paper.

## Who ends up in which profile? (covariates)

```{r covariates, cache = TRUE}
covs <- add_covariates(
  fit,
  ~ age + male + sh_experience + org_intolerance_sh + masc_job_context +
    anger + empathy + curb_expectancy,
  data = liang_ark_sim
)
summary(covs)
```

This is the three-step approach: the measurement model above is not
re-estimated when the covariates go in, and the classification error from
step three is corrected for. The reference class is class 1
("Supportive only"), the default; the coefficients above read as the
log-odds (and odds ratios) of landing in class 2 or 3 rather than class 1.
Men are considerably more likely than women to be classified into the
"Disengaged" profile rather than "Supportive only", and higher perceived
organizational intolerance of harassment and higher anger both shift the
odds toward the more active profiles.

These coefficients are approximate even as a reproduction of the published
study, because the simulated covariates were generated from
class-conditional marginal moments while the published logits are mutually
adjusted. Signs and rough magnitudes reproduce for the covariates that were
significant in the original; `age` and `masc_job_context` were not
significant there and their signs here are noise.

## What follows from the profile? (distal outcomes)

```{r outcomes, cache = TRUE}
summary(add_outcome(fit, liang_ark_sim$harasser_aggression,
                    outcome_type = "continuous"))
summary(add_outcome(fit, liang_ark_sim$target_gratitude,
                    outcome_type = "continuous"))
summary(add_outcome(fit, liang_ark_sim$third_party_elevation,
                    outcome_type = "continuous"))
```

Across all three outcomes the pattern is the same shape as the profiles
themselves: the "Disengaged" class scores lowest and "Broad responders"
scores highest, with "Supportive only" in between, and every pairwise
contrast is significant. Confronting or reporting a harasser draws more
aggression from them, but it is also what earns the target's gratitude and
what elevates third parties' view of the situation — the active profiles
pay a real cost and get a real benefit. These are BCH-corrected means, not
plain averages within the modal-assigned class, so the classification
error in step three is already accounted for.

## A closing check the reader could not do with real data

```{r recovery}
table(class_assignments(fit), liang_ark_sim$class_true)
```

About 85% of cases are assigned to the profile they were actually
simulated from. That is the realistic figure, not a shortfall: the
generating design deliberately includes genuinely ambiguous respondents
(the `boundary` column, ignored everywhere above) so that entropy lands
near the published value of about 0.90 rather than the artificially sharp
separation you get by simulating from a correctly specified model with no
overlap. A real analysis has no `class_true` to check against, which is
exactly what `classification_diagnostics()` is for.

```{r diagnostics}
classification_diagnostics(fit)
```

## References

Liang, Y., & Park, Y. (2025). A spectrum of bystander actions: Latent
profile analysis of sexual harassment intervention behavior at work.
*Journal of Applied Psychology*. Advance online publication.
<https://dx.doi.org/10.1037/apl0001280>.
