---
title: "Latent Transition Analysis: Movement Between Statuses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent Transition Analysis: Movement Between Statuses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

Latent transition analysis (LTA) is the longitudinal extension of LCA in
which cases can *move*: at each occasion every case occupies a latent
status, and the model estimates both what the statuses look like
(measurement) and the probabilities of moving between them from one
occasion to the next (transitions) (Collins & Lanza, 2010, ch. 7–8).

## The data: learning to read, kindergarten to first grade

`ecls_reading` holds 3,575 children from the Early Childhood Longitudinal
Study, Kindergarten Class of 1998–99 (ECLS-K), assessed four times — the
fall and spring of kindergarten and the fall and spring of first grade — on
whether they had mastered five reading skills: letter recognition,
beginning sounds, ending sounds, sight words and words in context. The
skills are acquired roughly in that order, which is what makes these data
a textbook case for a *stage-sequential* LTA (Kaplan, 2008): the latent
statuses are stages of learning to read, and children are expected to move
forward through them.

```{r data}
data(ecls_reading)
items <- ecls_reading[, -1]          # the 20 indicators; column 1 is poverty
round(colMeans(items), 2)
```

Mastery of every skill rises from occasion to occasion, and at any one
occasion the skills are ordered: almost everyone knows their letters by
the spring of kindergarten, while reading words in context is still rare a
year later. The indicator columns are already in the wide, occasion-major
layout `fit_lta()` expects — the five items at `t1`, then the five at
`t2`, and so on.

## Fitting the LTA

`fit_lta()` needs the number of statuses and how the columns divide into
occasions. Three statuses is the choice the published analyses of these
data settle on, and `compare_longitudinal()` can be used to check it. By
default the item-response probabilities are held equal across occasions
(`measurement_invariance = "full"`), which is what makes "the same status
at different times" a meaningful phrase:

```{r fit, cache = TRUE}
fit <- fit_lta(items, n_statuses = 3, times = 4, measurement = "binary",
               n_init = 50, random_state = 7)
fit
```

Latent transition models have many local maxima, and the `Best solution`
line — every one of the 50 starts reached this one — is the evidence that
this is not one of them. When that count is low, raise `n_init`; see
`vignette("estimation")`.

### What the statuses are

Status numbers are arbitrary labels. Read the item-response probabilities
before naming them:

```{r measurement}
measurement_summary(fit)
```

The three statuses are three stages of learning to read:

* **Low alphabet knowledge** — about half know their letters and almost
  nothing beyond;
* **Early word reading** — letters and beginning sounds mastered, ending
  sounds coming, sight words not yet;
* **Early reading comprehension** — everything mastered except, for about
  half of them, words in context.

The `At the boundary` note is worth reading rather than fearing. Nobody in
the low-alphabet stage reads sight words or words in context, and everyone
in the comprehension stage knows their letters: those probabilities are 0
and 1 because that is what the stages *are*, not because the estimation
went wrong. The note only matters if a standard error for one of those
cells is needed, and says what to do then.

`fit_lta()` labels the statuses from most to least prevalent at the first
occasion, which here coincides with the developmental order, so status 1 is
the earliest stage and status 3 the latest.

### Prevalences and transitions

The two core quantities of an LTA are the status prevalences at each
occasion and the transition matrices:

```{r extract}
status_prevalences(fit)
transition_matrix(fit)
```

Rows are where children start, columns where they end up, so the diagonal
holds the probability of staying put. Seven in ten enter kindergarten in
the low-alphabet stage; by the spring of first grade eight in ten are in
the comprehension stage. Movement is almost entirely forward — the cells
below the diagonal are all near zero — and the interval with the least
movement is the middle one, the summer between kindergarten and first grade, where
60% of the low-alphabet children and 84% of the early-word readers stay
where they were. `plot()` shows both views:

```{r plot, fig.height = 4}
labs <- c("Low alphabet", "Early word reading", "Comprehension")
plot(fit, type = "prevalence", status_labels = labs)
```

## Is the transition process the same at every interval?

`transition_invariance = "full"` forces one transition matrix for all
intervals; comparing it against the unrestricted model is a
likelihood-ratio test of a time-homogeneous process:

```{r invariance, cache = TRUE}
fit_hom <- fit_lta(items, n_statuses = 3, times = 4, measurement = "binary",
                   transition_invariance = "full", n_init = 50,
                   random_state = 7)
lr_test(fit_hom, fit)
```

It is rejected decisively. That is the summer: a school year moves children
forward at a rate a vacation does not, and one shared matrix cannot express
it.

## A stage-sequential restriction

If the stages are truly ordered, backward moves are impossible by design
and can be fixed at zero with `forbidden_transitions`, which also saves
their parameters:

```{r forward, cache = TRUE}
forward_only <- matrix(c(0, 0, 0,     # from Low alphabet: anything allowed
                         1, 0, 0,     # from Early word:   no return to Low
                         1, 1, 0),    # from Comprehension: no return at all
                       3, 3, byrow = TRUE)
fit_fwd <- fit_lta(items, n_statuses = 3, times = 4, measurement = "binary",
                   forbidden_transitions = forward_only, n_init = 50,
                   random_state = 7)
c(unrestricted = fit$metrics$bic, forward_only = fit_fwd$metrics$bic)
```

The two BICs are almost tied, and the unrestricted fit is 40 log-likelihood
units better for its nine extra parameters. The backward transitions it
estimates are tiny, but with a sample this size a handful of children who
look as though they slipped back is enough for the model to want them.
Whether those slips are real regression or measurement noise — a child who
knew ending sounds in the spring and missed the item in the fall — is a
question regular LTA cannot answer, because it has no way to represent a
stable difference between children that is not a stage. That is what a
random intercept adds; see `vignette("rilta")`.

## A mover-stayer model

A *mover-stayer* model (Vermunt, 2004) adds a latent class **above** the
chain: one group transitions freely (the movers) while the other is locked
in place (the stayers). It asks whether there is a genuinely immobile
group — here, children who do not progress at all across the two years:

```{r mover-stayer, cache = TRUE}
fit_ms <- fit_lta(items, n_statuses = 3, times = 4, measurement = "binary",
                  mover_stayer = TRUE, n_init = 20, random_state = 7,
                  max_iter = 2000)
c(single_chain = fit$metrics$bic, mover_stayer = fit_ms$metrics$bic)
```

The stayer class is small (about 6%) and buys too little likelihood for its
three parameters: BIC prefers the single chain. On these data nearly
everyone moves.

## Poverty as a covariate

Covariates can predict the status children start in
(`predictors_initial`) and the transitions they make
(`predictors_transition`). One slope per covariate is shared across origin
statuses within each interval (`transition_effects = "common"`), which is
the well-behaved specification; the alternative, a separate regression per
origin, is saturated and fails to converge usefully when transitions are
sparse.

```{r poverty, cache = TRUE}
fit_pov <- fit_lta(items, n_statuses = 3, times = 4, measurement = "binary",
                   predictors_initial = ecls_reading["poverty"],
                   predictors_transition = ecls_reading["poverty"],
                   n_init = 20, random_state = 7)
c(no_covariate = fit$metrics$bic, poverty = fit_pov$metrics$bic)
```

Poverty improves BIC by some 350 points. One thing to check before reading
the coefficients: once covariates index the statuses, `fit_lta()` no longer
relabels them by prevalence, so the numbering may differ from the fit
above. `status_prevalences()` says which is which —

```{r poverty-labels}
round(status_prevalences(fit_pov), 3)
```

— here status 3 is the low-alphabet stage (the one 69% start in), status 2
early word reading and status 1 comprehension. The coefficients are
multinomial logits against the last status, so with this labelling every
contrast is *relative to the low-alphabet stage*:

```{r poverty-coefs}
lta_covariate_summary(fit_pov)
```

Poverty lowers the odds of starting kindergarten beyond the low-alphabet
stage (OR 0.23 for early word reading, 0.12 for comprehension), and lowers
the odds of reaching comprehension in every interval. The effect is
sharpest over the summer, where a child in poverty has 0.40 times the odds
of reaching early word reading and 0.12 times the odds of reaching
comprehension, relative to staying in low alphabet. The `from:` terms are the origin-status
intercepts and are not of interest in themselves; the ones with enormous
standard errors belong to cells almost nobody occupies.

Whether the poverty effect on the *transitions* survives once stable
between-child differences in reading readiness are modelled directly is
the question `vignette("rilta")` takes up.

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent
transition analysis: With applications in the social, behavioral, and
health sciences*. Wiley.

Kaplan, D. (2008). An overview of Markov chain methods for the study of
stage-sequential developmental processes. *Developmental Psychology*,
*44*(2), 457–467. <https://doi.org/10.1037/0012-1649.44.2.457>

Vermunt, J. K. (2004). Mover-stayer models. In M. S. Lewis-Beck, A.
Bryman, & T. F. Liao (Eds.), *The SAGE encyclopedia of social science
research methods* (Vol. 3, p. 666). SAGE Publications.
<https://doi.org/10.4135/9781412950589.n583>
