---
title: "Latent Transition Analysis: Movement Between Statuses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Latent Transition Analysis: Movement Between Statuses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

Latent transition analysis (LTA) is the longitudinal extension of LCA in
which cases can *move*: at each occasion every case occupies a latent
status, and the model estimates both what the statuses look like
(measurement) and the probabilities of moving between them from one
occasion to the next (transitions) (Collins & Lanza, 2010, ch. 7–8).

## Simulated example: depression status across three assessments

We simulate 1,000 adults assessed three times on four binary symptom
indicators. The truth has two statuses — *Depressed* and *Not depressed* —
with most people staying where they are between assessments, and remission
(Depressed → Not depressed) more common than onset:

```{r simulate}
set.seed(2026)
n   <- 1000
rho <- c(Depressed = 0.85, `Not depressed` = 0.15)  # P(symptom | status)
tau <- matrix(c(0.80, 0.20,    # Depressed    -> (Depressed, Not)
                0.10, 0.90),   # Not depressed-> (Depressed, Not)
              2, 2, byrow = TRUE)

status <- matrix(0L, n, 3)
status[, 1] <- rbinom(n, 1, 0.35) + 1L           # 1 = Depressed, 2 = Not
for (t in 2:3)
  status[, t] <- ifelse(runif(n) < tau[status[, t - 1], 1], 1L, 2L)

sym <- matrix(0L, n, 12)
for (t in 1:3) for (j in 1:4)
  sym[, (t - 1) * 4 + j] <- rbinom(n, 1, rho[status[, t]])
colnames(sym) <- paste0("sym", 1:4, "_t", rep(1:3, each = 4))
```

## Fitting the LTA

`fit_lta()` needs the number of statuses and how the columns divide into
occasions. By default the item-response probabilities are held equal
across occasions (`measurement_invariance = "full"`), which is what makes
"the same status at different times" a meaningful phrase:

```{r fit}
fit <- fit_lta(sym, n_statuses = 2, times = 3, measurement = "binary",
               n_init = 20, random_state = 3)
fit
```

The two core quantities of an LTA are the status prevalences at each
occasion and the transition matrices:

```{r extract}
status_prevalences(fit)
transition_matrix(fit)
```

The recovered transitions sit close to the generating matrix: strong
diagonal (stability), with the Depressed → Not depressed cell larger than
its mirror image. `plot()` shows both:

```{r plot, fig.height = 4}
plot(fit, type = "prevalence", status_labels = c("Depressed", "Not depressed"))
```

## Is the transition process the same at every interval?

`transition_invariance = "full"` forces one transition matrix for all
intervals; comparing it against the unrestricted model is a likelihood-
ratio test of a time-homogeneous process:

```{r invariance}
fit_hom <- fit_lta(sym, n_statuses = 2, times = 3, measurement = "binary",
                   transition_invariance = "full", n_init = 20,
                   random_state = 3)
longitudinal_lrt(fit_hom, fit)
```

As simulated (one matrix generated both intervals), the restriction is not
rejected, and the more parsimonious homogeneous model is preferable.

## A mover-stayer model

A *mover-stayer* model (Vermunt, 2004) adds a latent
class **above** the chain: one group transitions freely (the movers) while
the other is locked in place (the stayers). It answers a question a single
chain cannot: is apparent stability just slow movement, or is there a
genuinely immobile group?

```{r mover-stayer, cache = TRUE}
fit_ms <- fit_lta(sym, n_statuses = 2, times = 3, measurement = "binary",
                  mover_stayer = TRUE, n_init = 20, random_state = 3,
                  max_iter = 2000)
fit_ms
```

On these data the mover-stayer model offers no improvement — the data were
generated from a single chain, and the fitted stayer class is essentially
redundant. On real data, comparing the two BICs (single chain vs.
mover-stayer) is the standard way to decide.

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent
transition analysis: With applications in the social, behavioral, and
health sciences*. Wiley.

Vermunt, J. K. (2004). Mover-stayer models. In M. S. Lewis-Beck, A.
Bryman, & T. F. Liao (Eds.), *The SAGE encyclopedia of social science
research methods* (Vol. 3, p. 666). SAGE Publications.
<https://doi.org/10.4135/9781412950589.n583>
