---
title: "Multiple-Group LCA: Do the Classes Differ by Grade?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple-Group LCA: Do the Classes Differ by Grade?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## The question

A latent class model assumes the classes mean the same thing to every case
in the data. When the data span groups — here, four grades of high school
— that assumption is worth checking rather than taking for granted, and it
splits into two separate questions. First, *measurement invariance*: do the
twelve health-risk items mean the same thing to a ninth-grader as to a
twelfth-grader, in the sense that the same class membership implies the
same item-response probabilities regardless of grade? Second, if
invariance holds, *prevalence*: are the risk classes more common in some
grades than others? The second question is only interpretable once the
first is settled — a prevalence comparison between groups whose items do
not function the same way is comparing different things.

This uses the same `yrbs2005` data as `vignette("survey_lca")`, but
unweighted, matching how the textbook this pipeline follows presents it;
see that vignette for the same items analysed with the sampling design in
play.

## Data

`grade` is a factor with levels 9 through 12, with no missing values in the
bundled sample:

```{r data}
data(yrbs2005)
items <- yrbs2005[, 3:14]
grade <- yrbs2005$grade
table(grade)
```

## The baseline: prevalences free, measurement invariant

```{r baseline, cache = TRUE}
set.seed(1)
m_free <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "prevalence",
                      n_init = 20, max_iter = 2000, n_steps = 1)
m_free
```

`group_effects = "prevalence"` pools the item-response probabilities
across grades — that pooling *is* the measurement-invariance assumption —
while letting each grade have its own class sizes. A four-group,
five-class fit on this many cases is materially more expensive per start
than the package's other vignettes; the solution above replicated across
6 of the 20 starts.

## The constrained model and the omnibus test

Fitting the same model with `group_effects = "none"` forces every grade to
share the same class sizes as well, which is the null hypothesis that
prevalence does not differ by grade at all:

```{r constrained, cache = TRUE}
set.seed(1)
m_none <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "none",
                      n_init = 5, max_iter = 2000, n_steps = 1)
lr_test(m_none, m_free)
```

`lr_test()` takes the restricted model first and the full model second —
reversing the order is an error. With four grades and five classes there
are (4 - 1) x (5 - 1) = 12 free prevalence contrasts between the two
models, and the printed `df` above confirms it. The restriction is
rejected: class prevalence does differ by grade. This reproduces the "all
five latent classes" comparison from the textbook's prevalence-differences
table.

Because `group = grade` is supplied, both fits above also report a
log-likelihood on the known-class scale, `metrics$ll_knownclass` — that is
the number that lines up with what other programs print for this kind of
model, as opposed to `metrics$ll`, which is on this package's own
1-step scale.

```{r ll-knownclass}
m_free$metrics$ll_knownclass
m_none$metrics$ll_knownclass
```

## Why the per-class tests are not here

The textbook's per-class comparisons each hold *one* class's marginal
probability exactly equal across grades while the other four float and
renormalise. That is a nonlinear constraint on the softmax class
probabilities, not a linear constraint on the underlying multinomial-logit
coefficients, and mixtureEM — like most implementations — constrains the
coefficients, not the probabilities directly. So that specific per-class
test is not reproduced here.

The linear alternative that *is* available answers a closely related
question: does grade shift the log-odds of landing in a given class
against the reference class at all? `add_covariates()` provides exactly
that, with an omnibus Wald test across all classes at once — but it needs
an unconditional step-1 fit to start from, and `m_free` does not qualify:
having been fit with `group = grade`, its classes already condition on
grade as a class-membership predictor, so grade cannot be added to it a
second time as a covariate. The fix is the same either way — fit the plain
(ungrouped) model and add grade afterwards:

```{r covariate-alt, cache = TRUE}
set.seed(1)
base_fit <- fit_mixture(items, n_classes = 5, measurement = "binary",
                        n_init = 20, max_iter = 2000)
covs <- add_covariates(base_fit, ~ grade, data = data.frame(grade = grade))
summary(covs)
```

## Testing measurement invariance itself

Section 3's pooling of the item probabilities across grades was an
assumption, not a given — it can be tested the same way the prevalence
question was. `group_effects = "both"` frees the item-response
probabilities by grade as well as the class sizes, and comparing it with
the invariant baseline via `lr_test()` tests exactly that assumption:

```{r invariance, cache = TRUE}
set.seed(1)
m_both <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "both",
                      n_init = 20, max_iter = 2000, n_steps = 1)
lr_test(m_free, m_both)
```

`m_both` carries a warning worth repeating rather than hiding: only 1 of
the 20 requested starts converged to the log-likelihood reported above,
so this solution has not been shown to be the global optimum the way
`m_free`'s was. 256 parameters spread across five classes and four
grades is a much harder surface to search than the models earlier in
this vignette, and `n_init = 20` is not enough to be confident of it here
— the package's own advice is to refit with `n_init = 100` before
reporting this specific comparison as a finding rather than an
illustration.

With N in the thousands, a likelihood-ratio test like this one also has
power to detect even a trivially small departure from invariance, so a
rejection on its own does not settle the question — it is worth reading
the information criteria alongside the test rather than the p-value in
isolation. Here the test rejects invariance decisively, which on its own
would be unsurprising at this sample size. But the two information
criteria disagree about which model is preferable: AIC favours the fully
group-varying model (it is smaller for `m_both`), while BIC — which
penalises the 180 extra parameters more heavily — favours the invariant
`m_free`. That split is itself informative: the departures from
invariance are real enough for the likelihood-ratio test to see them at
this N, but not so large that a criterion penalising complexity thinks
they are worth the extra parameters. A researcher would want to look at
*which* items differ by grade — and would want the better-replicated fit
above — before deciding whether to treat the invariant model as good
enough to interpret.

## Labelling

```{r label-check}
measurement_summary(m_free)
```

This fit is unweighted and grouped, and its classes do not read the same
way `vignette("survey_lca")`'s do — which is precisely the point of
comparing them, and why the five labels used there are not reused
unchanged here. That fit had one class high on essentially every item;
this one splits that pattern in two: class 4 is where the hard-drug items
(`cocaine_ever`, `glue_ever`, `meth_ever`, `ecstasy_ever`) and the smoking
items peak, while class 5 is instead the class where the two sexual-risk
items (`sex_before_13`, `sex_4plus_partners`) are far higher than in any
other class, at moderate levels of everything else. There is no single
"high risk across everything" class in this solution. Reading the table by
the same descriptive rule as before — lowest on nearly everything, highest
on a specific cluster of items relative to the rest — gives:

```{r plot}
plot(m_free,
     class_labels = c("Low risk", "Alcohol-focused", "Early experimenters",
                      "Hard drug users", "High sexual risk"),
     main = "Health-risk behavior classes by grade (YRBS 2005, unweighted)")
```

Class 2 stands out on `drove_drinking` and `binge_drink_30d` specifically;
class 3 stands out on the two early-onset items, `smoked_before_13` and
`first_drink_before_13`, more than on the items that come later in
adolescence. Both read the same way they did in the weighted fit, which is
why "Alcohol-focused" and "Early experimenters" carry over; "Hard drug
users" and "High sexual risk" replace the single "High risk" class because
the two clusters of items no longer travel together in one class here.

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent transition
analysis: With applications in the social, behavioral, and health
sciences*. Wiley.
