---
title: "Multiple-Group LCA: Do the Classes Differ by Grade?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple-Group LCA: Do the Classes Differ by Grade?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## The question

A latent class model assumes the classes mean the same thing to every case
in the data. When the data span groups — here, four grades of high school
— that assumption is worth checking rather than taking for granted, and it
splits into two separate questions. First, *measurement invariance*: do the
twelve health-risk items mean the same thing to a ninth-grader as to a
twelfth-grader, in the sense that the same class membership implies the
same item-response probabilities regardless of grade? Second, if
invariance holds, *prevalence*: are the risk classes more common in some
grades than others? The second question is only interpretable once the
first is settled — a prevalence comparison between groups whose items do
not function the same way is comparing different things.

This uses the same `yrbs2005` data as `vignette("survey_lca")`, but
unweighted, matching how the textbook this pipeline follows presents it;
see that vignette for the same items analysed with the sampling design in
play.

## Data

`grade` is a factor with levels 9 through 12, with no missing values in the
bundled sample:

```{r data}
data(yrbs2005)
items <- yrbs2005[, 3:14]
grade <- yrbs2005$grade
table(grade)
```

## The baseline: prevalences free, measurement invariant

```{r baseline, cache = TRUE}
set.seed(1)
m_free <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "prevalence",
                      n_init = 20, max_iter = 2000, n_steps = 1)
m_free
```

`group_effects = "prevalence"` pools the item-response probabilities
across grades — that pooling *is* the measurement-invariance assumption —
while letting each grade have its own class sizes. A four-group,
five-class fit on this many cases is materially more expensive per start
than the package's other vignettes; the solution above replicated across
6 of the 20 starts.

## The constrained model and the omnibus test

Fitting the same model with `group_effects = "none"` forces every grade to
share the same class sizes as well, which is the null hypothesis that
prevalence does not differ by grade at all:

```{r constrained, cache = TRUE}
set.seed(1)
m_none <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "none",
                      n_init = 5, max_iter = 2000, n_steps = 1)
lr_test(m_none, m_free)
```

`lr_test()` takes the restricted model first and the full model second —
reversing the order is an error. With four grades and five classes there
are (4 - 1) x (5 - 1) = 12 free prevalence contrasts between the two
models, and the printed `df` above confirms it. The restriction is
rejected: class prevalence does differ by grade. This reproduces the "all
five latent classes" comparison from the textbook's prevalence-differences
table.

Because `group = grade` is supplied, both fits above also report a
log-likelihood on the known-class scale, `metrics$ll_knownclass` — that is
the number that lines up with what other programs print for this kind of
model, as opposed to `metrics$ll`, which is on this package's own
1-step scale.

```{r ll-knownclass}
m_free$metrics$ll_knownclass
m_none$metrics$ll_knownclass
```

## Per-class tests: does *this* class's prevalence differ by grade?

The omnibus test above says prevalence differs by grade somewhere among the
five classes, but the textbook goes further and asks the question
separately for each class: hold Low Risk's share fixed across grades while
the other four float and renormalise, refit, and see how much likelihood
that costs. Repeated five times, this reproduces the chapter's Table 5.24.
`group_prevalence_equal` fits exactly this restriction — pass it the index
of the class to freeze and it holds that column of the per-group class
probabilities to one shared value while the rest adjust freely.

There is a wrinkle a straightforward call does not handle. A mixture
model's classes are not identified independently across separate fits:
nothing pins "class 1" to the same substantive group of respondents from
one `fit_mixture()` call to the next, so a search run five times, once per
requested class, is free to relabel *which* real class lands in whichever
slot was asked for. Left to its own devices the optimiser takes the escape
hatch every time — it reports whichever class is *cheapest* to freeze,
regardless of which one was requested, because that fit has the higher
likelihood and wins the random-restart comparison. Running
`group_prevalence_equal` directly, five times, mostly reproduces the same
answer (the cost of freezing High Risk, the cheapest class) instead of five
different ones.

The fix is to anchor each fit to `m_free`'s own solution rather than
searching from scratch. `class_sizes()`'s `by_group` attribute gives the
per-grade class probabilities `m_free` already settled on; seeding a fit at
those values, plus `m_free`'s own item-response probabilities, starts EM
in the one basin where "class k" already means the same thing it means in
`m_free`. `fit_mixture_internal()` (the lower-level engine behind
`fit_mixture()`) takes a `warm_start` function for exactly this, and
`n_init = 0` tells it to refine only from that seed rather than compete it
against random restarts:

```{r per-class-setup, cache = TRUE}
by_group <- attr(class_sizes(m_free), "by_group")
n_grades <- nlevels(grade)
n_classes <- 5
gamma0 <- matrix(0, n_grades, n_classes)
for (i in seq_len(nrow(by_group))) {
  gr <- match(by_group$group[i], levels(grade))
  gamma0[gr, by_group$class[i]] <- by_group$proportion[i]
}

anchor_at_m_free <- function(model_state, X, Y) {
  model_state$weights <- m_free$weights
  model_state$mm <- m_free$mm
  model_state$sm$parameters$gamma <- gamma0
  model_state
}

class_labels <- c("Low Risk", "Binge Drinkers", "Early Experimenters",
                  "High Risk", "Sexual Risk-Takers")
freeze_class <- function(k) {
  fit_mixture_internal(X = as.matrix(items), Y = as.integer(grade),
                       n_components = n_classes, measurement = "binary",
                       structural = "group_prevalence", n_steps = 1,
                       n_init = 0, max_iter = 2000,
                       warm_start = anchor_at_m_free,
                       n_groups = n_grades, frozen = k)
}
```

```{r per-class-tests, cache = TRUE}
per_class <- lapply(1:5, function(k) lr_test(freeze_class(k), m_free))
data.frame(class = class_labels,
          dG2 = vapply(per_class, function(t) t$statistic, numeric(1)),
          df = vapply(per_class, function(t) t$df, numeric(1)),
          p_value = vapply(per_class, function(t) t$p_value, numeric(1)))
```

Each row matches the textbook's Table 5.24 to about one decimal (61.3,
468.5, 176.6, 0.4 and 5.7, in the same class order as above), and
the pattern makes substantive sense against the item plot above: Binge
Drinkers, the class whose share changes most across grades in that plot, is
the most expensive to force flat by a wide margin, while High Risk barely
moves across grades and costs almost nothing to freeze. The all-classes
restriction from the previous section (dG2 = 624.8, df = 12, against the
book's 625.5) is the sum of a similar idea applied to every class's share
at once, not to each one separately.

## Testing measurement invariance itself

Section 3's pooling of the item probabilities across grades was an
assumption, not a given — it can be tested the same way the prevalence
question was. `group_effects = "both"` frees the item-response
probabilities by grade as well as the class sizes, and comparing it with
the invariant baseline via `lr_test()` tests exactly that assumption:

```{r invariance, cache = TRUE}
set.seed(1)
m_both <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "both",
                      n_init = 20, max_iter = 2000, n_steps = 1)
lr_test(m_free, m_both)
```

`m_both` carries a warning worth repeating rather than hiding: only 1 of
the 20 requested starts converged to the log-likelihood reported above,
so this solution has not been shown to be the global optimum the way
`m_free`'s was. 256 parameters spread across five classes and four
grades is a much harder surface to search than the models earlier in
this vignette, and `n_init = 20` is not enough to be confident of it here
— the package's own advice is to refit with `n_init = 100` before
reporting this specific comparison as a finding rather than an
illustration.

With N in the thousands, a likelihood-ratio test like this one also has
power to detect even a trivially small departure from invariance, so a
rejection on its own does not settle the question — it is worth reading
the information criteria alongside the test rather than the p-value in
isolation. Here the test rejects invariance decisively, which on its own
would be unsurprising at this sample size. But the two information
criteria disagree about which model is preferable: AIC favours the fully
group-varying model (it is smaller for `m_both`), while BIC — which
penalises the 180 extra parameters more heavily — favours the invariant
`m_free`. That split is itself informative: the departures from
invariance are real enough for the likelihood-ratio test to see them at
this N, but not so large that a criterion penalising complexity thinks
they are worth the extra parameters. A researcher would want to look at
*which* items differ by grade — and would want the better-replicated fit
above — before deciding whether to treat the invariant model as good
enough to interpret.

## Labelling

```{r label-check}
measurement_summary(m_free)
```

This fit is unweighted and grouped, and its classes do not read the same
way `vignette("survey_lca")`'s do — which is precisely the point of
comparing them. That fit had one class high on essentially every item;
this one splits that pattern in two: class 4 is where the hard-drug items
(`cocaine_ever`, `glue_ever`, `meth_ever`, `ecstasy_ever`) and the smoking
items peak, while class 5 is instead the class where the two sexual-risk
items (`sex_before_13`, `sex_4plus_partners`) are far higher than in any
other class, at moderate levels of everything else. There is no single
"high risk across everything" class in this solution. In size order, the
five classes are **Low Risk, Binge Drinkers, Early Experimenters, High
Risk,** and **Sexual Risk-Takers**:

```{r plot}
plot(m_free,
     class_labels = c("Low Risk", "Binge Drinkers", "Early Experimenters",
                      "High Risk", "Sexual Risk-Takers"),
     main = "Health-risk behavior classes by grade (YRBS 2005, unweighted)")
```

Class 2 (Binge Drinkers) stands out on `drove_drinking` and
`binge_drink_30d` specifically; class 3 (Early Experimenters) stands out
on the two early-onset items, `smoked_before_13` and
`first_drink_before_13`, more than on the items that come later in
adolescence. High Risk and Sexual Risk-Takers replace the single "High
risk" class from the weighted fit because the hard-drug cluster and the
sexual-risk cluster no longer travel together in one class here.

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent transition
analysis: With applications in the social, behavioral, and health
sciences*. Wiley.
