---
title: "Multiple-Group LCA: Do the Classes Differ by Grade?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple-Group LCA: Do the Classes Differ by Grade?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## The question

A latent class model assumes the classes mean the same thing to every case in
the data. When the data span groups — here, four grades of high school — that
assumption is worth checking rather than taking for granted, and it splits into
two questions that have to be asked in that order. First, *measurement
invariance*: do the twelve health-risk items mean the same thing to a
ninth-grader as to a twelfth-grader, in the sense that the same class membership
implies the same item-response probabilities regardless of grade? Second, and
only once the first is settled, *prevalence*: are the risk classes more common
in some grades than others? A prevalence comparison between groups whose items
do not function the same way is comparing different things.

Clogg and Goodman (1985) introduced the simultaneous — multiple-group — latent
structure model, and set out the three models as a ladder. Their vocabulary is
worth having, because it is the vocabulary the literature uses:

| The model | `group_effects` | Clogg and Goodman's name |
|---|---|---|
| items and class sizes both free by grade | `"both"` | heterogeneous, unrestricted |
| items pooled, class sizes free by grade | `"prevalence"` | partially homogeneous |
| items and class sizes both pooled | `"none"` | homogeneous |

The middle rung is the one that carries the invariance assumption: in it, as
Kankaraš, Moors and Vermunt (2011) put it, "the relationships between indicator
items and the latent variable are identical across groups so that the class
memberships have the same meaning in all groups." So the pipeline below fits all
three models, tests the middle rung against the top one first, and only then
tests the bottom rung against the middle.

The worked example is Collins and Lanza's (2010, ch. 5). It uses the same
`yrbs2005` data as `vignette("survey_lca")`, but unweighted, matching how they
present it; see that vignette for the same items analysed with the sampling
design in play.

## Data

`grade` is a factor with levels 9 through 12, with no missing values in the
bundled sample:

```{r data}
data(yrbs2005)
items <- yrbs2005[, 3:14]
grade <- yrbs2005$grade
table(grade)
```

## Step 1: do the items mean the same thing in every grade?

The invariant model first. `group_effects = "prevalence"` pools the
item-response probabilities across grades — that pooling *is* the
measurement-invariance assumption — while letting each grade have its own class
sizes:

```{r baseline, cache = TRUE}
set.seed(1)
m_free <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "prevalence",
                      n_init = 20, max_iter = 2000, n_steps = 1)
m_free
```

A four-group, five-class fit on this many cases is materially more expensive per
start than the package's other vignettes; the solution above replicated across
6 of the 20 starts.

`group_effects = "both"` frees the item-response probabilities by grade as well
as the class sizes, and comparing the two by likelihood ratio tests exactly the
assumption the invariant model is making:

```{r invariance, cache = TRUE}
set.seed(1)
m_both <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "both",
                      n_init = 20, max_iter = 2000, n_steps = 1)
lr_test(m_free, m_both)
```

`lr_test()` takes the restricted model first and the full model second —
reversing the order is an error.

`m_both` carries a warning worth repeating rather than hiding: only 1 of the 20
requested starts converged to the log-likelihood reported above, so this solution
has not been shown to be the global optimum the way `m_free`'s was. 256
parameters spread across five classes and four grades is a much harder surface to
search than the models later in this vignette, and `n_init = 20` is not enough to
be confident of it here — the package's own advice is to refit with
`n_init = 100` before reporting this specific comparison as a finding rather than
an illustration.

With N in the thousands, a likelihood-ratio test like this one also has power to
detect even a trivially small departure from invariance, so a rejection on its
own does not settle the question. Kankaraš, Moors and Vermunt (2011) make the
point directly: "when sample sizes are large, likelihood-ratio tests tend to be
too conservative, indicating misfit even for minimal differences between two
models", and "since they also control for sample size, BIC and CAIC are preferred
fit statistics in situations when sample size is large." Here the test rejects
invariance decisively, which on its own would be unsurprising at this sample
size. The two information criteria disagree about which model is preferable: AIC
favours the fully group-varying model (it is smaller for `m_both`), while BIC
favours the invariant `m_free`. That split is not, on its own, evidence that the
departure from invariance is small. Finch (2015) simulates designs comparable to
this one in class count and group count and finds that information criteria have
very low power to detect measurement noninvariance in multiple-group latent class
models — a criterion preferring the invariant model is therefore weak evidence
for invariance, not evidence that the true departure is minor. The reason to
carry `m_free` forward is the one already given above: at this N the
likelihood-ratio test has power against departures too small to matter for
interpretation, not that BIC's preference confirms them small.

The four grades are also not equal in size, and Finch's simulations show that
unequal groups both reduce power to detect real noninvariance and inflate the
false-positive rate for a prevalence comparison made afterward — a caution worth
carrying into Step 2 below. Collins and Lanza's own sequencing rule, which
Finch's design follows, is not to test prevalence once measurement invariance
has been rejected; this vignette tests it anyway, on the view that the rejection
here reflects the sample size rather than a difference large enough to make the
prevalence comparison uninterpretable. Step 2's result should be read with that
qualification attached, not as a comparison the literature treats as routine
once Step 1 has failed.

`m_free` is therefore the model the rest of this vignette carries forward. A
researcher would want to look at *which* items differ by grade — and would want
the better-replicated fit — before treating the invariant model as settled rather
than as good enough to interpret.

## The five classes

```{r label-check}
measurement_summary(m_free)
```

This fit is unweighted and grouped, and its classes do not read the same way
`vignette("survey_lca")`'s do — which is precisely the point of comparing them.
That fit had one class high on essentially every item; this one splits that
pattern in two: class 4 is where the hard-drug items (`cocaine_ever`,
`glue_ever`, `meth_ever`, `ecstasy_ever`) and the smoking items peak, while
class 5 is instead the class where the two sexual-risk items (`sex_before_13`,
`sex_4plus_partners`) are far higher than in any other class, at moderate levels
of everything else. There is no single "high risk across everything" class in
this solution. In size order, the five classes are **Low Risk, Binge Drinkers,
Early Experimenters, High Risk,** and **Sexual Risk-Takers**:

```{r plot}
class_labels <- c("Low Risk", "Binge Drinkers", "Early Experimenters",
                  "High Risk", "Sexual Risk-Takers")
plot(m_free, class_labels = class_labels,
     main = "Health-risk behavior classes by grade\n(YRBS 2005, unweighted)")
```

Class 2 (Binge Drinkers) stands out on `drove_drinking` and `binge_drink_30d`
specifically; class 3 (Early Experimenters) stands out on the two early-onset
items, `smoked_before_13` and `first_drink_before_13`, more than on the items
that come later in adolescence. High Risk and Sexual Risk-Takers replace the
single "High risk" class from the weighted fit because the hard-drug cluster and
the sexual-risk cluster no longer travel together in one class here.

## Step 2: are the classes more common in some grades than others?

Fitting the same model with `group_effects = "none"` forces every grade to
share the same class sizes as well, which is the null hypothesis that
prevalence does not differ by grade at all — Clogg and Goodman's homogeneous
model:

```{r constrained, cache = TRUE}
set.seed(1)
m_none <- fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "none",
                      n_init = 5, max_iter = 2000, n_steps = 1)
lr_test(m_none, m_free)
```

With four grades and five classes there are (4 - 1) x (5 - 1) = 12 free
prevalence contrasts between the two models, and the printed `df` above confirms
it. The restriction is rejected: class prevalence does differ by grade. This
reproduces the "all five latent classes" comparison from Collins and Lanza's
prevalence-differences table.

Because `group = grade` is supplied, `m_free` and `m_none` also report a
log-likelihood on the known-class scale, `metrics$ll_knownclass` — that is the
number that lines up with what other programs print for this kind of model, as
opposed to `metrics$ll`, which is on this package's own 1-step scale.

```{r ll-knownclass}
m_free$metrics$ll_knownclass
m_none$metrics$ll_knownclass
```

## Step 3: which class's prevalence differs by grade?

The omnibus test says prevalence differs by grade somewhere among the five
classes. Collins and Lanza go further and ask the question separately for each
class: hold Low Risk's share fixed across grades while the other four float and
renormalise, refit, and see how much likelihood that costs. Repeated five times,
this reproduces their Table 5.24. `group_prevalence_equal` fits exactly this
restriction — pass it the index of the class to freeze and it holds that column
of the per-grade class probabilities to one shared value while the rest adjust
freely.

Two things about the restriction are worth knowing before reading the table.

**It is tested by likelihood ratio, not by a Wald test on the class-membership
regression's group coefficients, and that is deliberate.** A Wald test on a group
dummy asks whether one class's log-odds relative to the reference class departs
from the reference group's — a comparison relative to the group average, not "is
this class's share literally the same in every grade" — and the two questions can
disagree. Clogg and Goodman single this restriction out as one needing "special
treatment": for a restriction of exactly this form, they note, "the group sizes
must be taken into account in order to obtain the correct restrictions and the
appropriate likelihood equations." `lr_test()` against the unrestricted fit is
the exact test of the question actually being asked.

**Each restricted fit has to be anchored to the unrestricted solution, so the
random restarts are switched off.** A mixture model's classes are named only by
their own parameters, and "class *k* has the same prevalence in every grade" is a
restriction on a *named* class: permute the labels and the frozen slot lands on a
different substantive class at exactly the same likelihood. The restricted
model's *global* maximum is therefore one and the same number whichever class is
named — whichever class is cheapest to freeze — and a search reports that number
every time, for every k. What the question asks for is the maximum near the
unrestricted solution, where class *k* still means what it means in `m_free`.
`start_from = m_free` supplies exactly that: it seeds each fit at `m_free`'s own
item-response probabilities and per-grade class sizes and runs no other start.
This is the one place in the package where a starting value replaces the search
rather than joining it, and other latent-class software arrives at the same
arrangement from the other side — the restriction is expressed there by supplying
the unrestricted solution as starting values, and supplying starting values makes
the random-start option unavailable in the same call.

```{r per-class, cache = TRUE}
per_class <- lapply(1:5, function(k)
  lr_test(fit_mixture(items, n_classes = 5, measurement = "binary",
                      group = grade, group_effects = "prevalence",
                      group_prevalence_equal = k, start_from = m_free,
                      n_steps = 1, max_iter = 2000),
          m_free))
data.frame(class = class_labels,
           dG2 = vapply(per_class, function(t) t$statistic, numeric(1)),
           df = vapply(per_class, function(t) t$df, numeric(1)),
           p_value = vapply(per_class, function(t) t$p_value, numeric(1)))
```

Each row matches Collins and Lanza's Table 5.24 to about one decimal (61.3,
468.5, 176.6, 0.4 and 5.7, in the same class order as above), and the pattern
makes substantive sense against the item plot above: Binge Drinkers, the class
whose share changes most across grades in that plot, is the most expensive to
force flat by a wide margin, while High Risk barely moves across grades and costs
almost nothing to freeze. The all-classes restriction from the previous section
(dG2 = 624.8, df = 12, against the book's 625.5) is the sum of a similar idea
applied to every class's share at once, not to each one separately.

## References

Clogg, C. C., & Goodman, L. A. (1985). Simultaneous latent structure analysis in
several groups. *Sociological Methodology*, *15*, 81-110.

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent transition
analysis: With applications in the social, behavioral, and health sciences*.
Wiley.

Finch, W. H. (2015). A comparison of methods for assessing measurement
invariance in multiple group latent class analysis. *Frontiers in Psychology*,
*6*, 1400.

Kankaraš, M., Moors, G., & Vermunt, J. K. (2011). Testing for measurement
invariance with latent class analysis. In E. Davidov, P. Schmidt, & J. Billiet
(Eds.), *Cross-cultural analysis: Methods and applications* (pp. 359-384).
Routledge.
