---
title: "Random-Intercept LTA: Separating Traits from Transitions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random-Intercept LTA: Separating Traits from Transitions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## The problem with ordinary LTA

Latent transition analysis puts every person in a latent status at each
occasion and estimates how they move between statuses. It assumes that,
within a status, people are exchangeable: two children in the *early word
reading* stage are equally likely to know their ending sounds, and any
difference between them is noise.

That assumption is usually false. People differ in stable, trait-like
ways that are not stages at all: a child with a generally high propensity
to endorse the items will look "further along" at every occasion, and an
ordinary LTA has only one way to represent that — as a status. The
consequences are systematic (Muthén & Asparouhov, 2022):

* **Stability is overstated.** A child who scores high at every occasion
  because of a stable trait is recorded as *staying* in an advanced
  status, so the diagonal of the transition matrix inflates.
* **Statuses absorb the trait.** Item-response probabilities are pushed
  towards 0 and 1, statuses look more distinct than they are, and BIC can
  ask for extra statuses that are really levels of the trait.
* **Covariate effects are misattributed.** A covariate that predicts the
  trait shows up as predicting transitions.

Random-intercept LTA (RI-LTA) adds a person-level factor to the
measurement model — a random intercept shared by every indicator at every
occasion — so that stable between-person differences have somewhere to go
other than the statuses. What is left in the transition matrices is then
*within-person* change.

This vignette walks through the pipeline an applied analysis needs:
ordinary LTA as a baseline, the random intercept, model comparison, and
covariates. It uses the reading-proficiency data introduced in
`vignette("lta")`; read that first for what the data are and how the
ordinary model is interpreted.

## Setup

```{r data}
data(ecls_reading)
items <- ecls_reading[, -1]
```

Two arguments appear on every fit below. `fit_lta()`'s defaults put a
light prior on the status and transition probabilities (`smoothing = 1`)
and on the item probabilities (`bayes_constants`), which keeps estimates
off the boundary and is the right default for exploratory work. Here
they are switched off so the fits are plain maximum likelihood, which is
what the published analyses of these data report and what the numbers
below can be checked against. On a sample this size the difference is
about one log-likelihood unit and no change in interpretation.

```{r ml}
ml <- list(smoothing = 0, bayes_constants = list(categorical = 0))
```

Every fit is also run at `standard_errors = FALSE` until the final model,
because the standard errors are the slow part and are not needed to
compare models.

## Step 1: the ordinary LTA

```{r lta, cache = TRUE}
fit_lta0 <- fit_lta(items, n_statuses = 3, times = 4, n_init = 50,
                    random_state = 7, standard_errors = FALSE,
                    smoothing = ml$smoothing, bayes_constants = ml$bayes_constants)
fit_lta0
```

Three stages of learning to read — *low alphabet knowledge*, *early word
reading*, *early reading comprehension* — with 69% of children in the
first at the start of kindergarten and 81% in the third by the end of
first grade. (The item probabilities, and the note about the ones at the
boundary, are in `vignette("lta")`.) Keep the transition matrices in mind:

```{r lta-tau}
transition_matrix(fit_lta0)
```

## Step 2: add a continuous random intercept

`random_intercept = "continuous"` adds a normally distributed factor with
one loading per indicator, held equal across occasions — if the loadings
changed over time, the factor would not be a stable trait. The factor is
integrated out numerically, and `n_quadrature` sets how finely: 30 nodes
is the setting the published analyses of these data used, and is checked
below.

```{r rilta, cache = TRUE}
fit_ri <- fit_lta(items, n_statuses = 3, times = 4, n_init = 20,
                  random_intercept = "continuous", n_quadrature = 30,
                  random_state = 7, standard_errors = FALSE,
                  smoothing = ml$smoothing, bayes_constants = ml$bayes_constants)
fit_ri
```

Two things before the numbers. The warning is the same substantive zero
the ordinary model has — nobody in the low-alphabet stage reads words in
context — now stated on the logit scale the random-intercept model
works on, where a probability of zero is a logit of minus infinity. It is
worth reading once; it recurs on every fit below for the same reason and
is silenced from here on. And `Best solution: found by 3 of 3 starts that
ran to convergence (of 20 requested)` is how the random-intercept search
reports: the twenty starts are ranked cheaply and only the best few are
run to convergence, so replication is counted among those.

```{r quiet}
knitr::opts_chunk$set(warning = FALSE)
```

The fit line to read is the log-likelihood: −20,329 against −21,793, for
five more parameters (the loadings). A BIC drop of almost three thousand
is not a subtle improvement. **Do not turn this into a likelihood-ratio
test**: the ordinary LTA is the random-intercept model with every loading
at zero, which is on the boundary of the parameter space, so the usual
chi-squared reference does not apply. Compare by BIC.

### The loadings

```{r loadings}
random_intercept_loadings(fit_ri)
```

Every loading is large — on the logit scale, a child one standard
deviation above average on the trait has odds of mastering sight words
`exp(3.75) ≈ 42` times those of an average child *in the same status*.
The trait is best read as reading readiness: a stable, time-invariant
dimension that the ordinary model had been squeezing into the stages.

### What changes in the statuses

```{r measurement}
measurement_summary(fit_ri)
```

Compared with the ordinary LTA, the item probabilities move away from 0
and 1: in the low-alphabet stage 63% now know their letters (not 51%)
and 30% their beginning sounds (not 7%), and in the comprehension stage
sight words are at 81% rather than 98%. The statuses are less extreme
because the extremity was partly the trait. The stages keep their meaning
and their order.

### What changes in the prevalences and transitions

```{r prevalences}
round(status_prevalences(fit_lta0), 3)
round(status_prevalences(fit_ri), 3)
```

The picture of the cohort changes considerably. Under the random-intercept
model 95% of children start in the low-alphabet stage (not 69%), 82%
are in early word reading by the spring of kindergarten (not 64%) and 97%
reach comprehension by the end of first grade (not 81%). The ordinary
model had spread children across statuses at every occasion partly on the
basis of a trait; with the trait removed, nearly all children are found
to pass through the same sequence, just at different times.

```{r transitions}
transition_matrix(fit_ri)
```

The transition matrices now describe a cohort moving almost as one: 82%
leave the low-alphabet stage over the first half of kindergarten, and
essentially everyone in early word reading moves up to comprehension over
the second half of first grade (98%, against 86% in the ordinary model).
The most common paths through the four occasions make the same point:

```{r paths}
paths <- function(fit, n = 3)
  head(transition_patterns(fit)[order(-transition_patterns(fit)$proportion), ], n)
paths(fit_lta0)
paths(fit_ri)
```

In the ordinary LTA the most common path accounts for a third of the
children; in the random-intercept model the path 1 → 2 → 2 → 3 — low
alphabet in the fall of kindergarten, early word reading through the
following year, comprehension by the end of first grade — accounts for
seven in ten.

### Is the quadrature fine enough?

The factor is integrated numerically, and too coarse a grid biases the
log-likelihood. `refine_from` continues the converged fit on a finer grid
without repeating the random-start search, so the check is cheap:

```{r quadrature, cache = TRUE}
fit_ri60 <- fit_lta(items, n_statuses = 3, times = 4,
                    random_intercept = "continuous", n_quadrature = 60,
                    refine_from = fit_ri, standard_errors = FALSE,
                    smoothing = ml$smoothing, bayes_constants = ml$bayes_constants)
c(nodes_30 = fit_ri$loglik, nodes_60 = fit_ri60$loglik)
round(cbind(nodes_30 = random_intercept_loadings(fit_ri)$loading,
            nodes_60 = random_intercept_loadings(fit_ri60)$loading), 3)
```

Doubling the grid moves the log-likelihood by less than 0.2 and the
loadings by less than 0.01 — nothing that touches a conclusion. The
log-likelihood is not monotone in the node count (see `?fit_lta`), so
check over a range rather than one step up when the fit's thresholds are
extreme.

## Step 3: a binary random intercept, and the model comparison

The continuous factor says people vary along a smooth dimension. The
alternative is two latent *types* of person — a binary random intercept,
`random_intercept = "binary"` — which asks a different question and is
compared by BIC like everything else:

```{r rilta-binary, cache = TRUE}
fit_ri2 <- fit_lta(items, n_statuses = 3, times = 4, n_init = 20,
                   random_intercept = "binary",
                   random_state = 7, standard_errors = FALSE,
                   smoothing = ml$smoothing, bayes_constants = ml$bayes_constants)
```

```{r comparison}
fits <- list(`Regular LTA` = fit_lta0,
             `RI-LTA, binary` = fit_ri2,
             `RI-LTA, continuous` = fit_ri)
data.frame(parameters = sapply(fits, `[[`, "n_params"),
           loglik     = round(sapply(fits, `[[`, "loglik"), 1),
           BIC        = round(sapply(fits, function(f) f$metrics$bic), 1))
```

Both random intercepts improve enormously on the ordinary LTA, and the
continuous one is far better than the binary one. That is the usual result
with binary indicators (Muthén & Asparouhov find the same on their own
example): a two-type intercept is a coarse approximation to a trait that
varies continuously. Write it up as a finding — "in these data the
continuous random-intercept model fits best" — not as a rule.

## Step 4: covariates

A covariate can act in three places, and RI-LTA is what lets them be told
apart: on the stable trait (`predictors_random_intercept`), on the status
children start in (`predictors_initial`) and on the transitions
(`predictors_transition`). These are different questions. Poverty
predicting the trait means poor children have lower reading readiness
throughout; poverty predicting transitions means they advance more slowly
*given* their readiness.

Random-start searches are slow once the factor is regressed on
covariates, and unnecessary: the unconditional fit is already the right
neighbourhood. `refine_from` continues from it in a single run.

```{r pov-ri, cache = TRUE}
pov <- ecls_reading["poverty"]
fit_pov_ri <- fit_lta(items, n_statuses = 3, times = 4,
                      random_intercept = "continuous", n_quadrature = 30,
                      predictors_random_intercept = pov,
                      refine_from = fit_ri, standard_errors = FALSE,
                      smoothing = ml$smoothing, bayes_constants = ml$bayes_constants)
fit_pov_tr <- fit_lta(items, n_statuses = 3, times = 4,
                      random_intercept = "continuous", n_quadrature = 30,
                      predictors_random_intercept = pov,
                      predictors_transition = pov,
                      refine_from = fit_pov_ri,
                      smoothing = ml$smoothing, bayes_constants = ml$bayes_constants)
```

```{r pov-comparison}
fits <- list(`RI-LTA` = fit_ri,
             `+ poverty -> trait` = fit_pov_ri,
             `+ poverty -> trait and transitions` = fit_pov_tr)
data.frame(parameters = sapply(fits, `[[`, "n_params"),
           loglik     = round(sapply(fits, `[[`, "loglik"), 1),
           BIC        = round(sapply(fits, function(f) f$metrics$bic), 1))
```

One parameter — poverty on the trait — buys 200 log-likelihood units.
Letting poverty also shift the transitions (six more parameters) buys
twenty-one more: a likelihood-ratio test of the two, which is legitimate
here because both models have the random intercept, rejects the
restriction decisively, while BIC prefers the smaller model. The two
criteria disagree because they weigh parsimony differently, and a
write-up should say so rather than pick the one it likes. Compare either
with `vignette("lta")`, where poverty on the initial status and the
transitions of the *ordinary* LTA was worth 210 units: most of that
effect was the trait all along.

```{r pov-test}
lr_test(fit_pov_ri, fit_pov_tr)
```

A fit continued with `refine_from` keeps its donor's status labels, so
status 1 is still the low-alphabet stage and status 3 comprehension —
check with `status_prevalences()` when in doubt — and every transition
contrast below is relative to *reaching comprehension*:

```{r pov-coefs}
lta_covariate_summary(fit_pov_tr)
```

Read the sign of the trait coefficient against the loadings, which are
positive: poverty lowers reading readiness by about 0.9 standard
deviations of the factor, a large and precisely estimated effect. With
that in the model the transition effects that remain are the ones
*given* readiness: over the summer a child in poverty has six times the
odds of staying in low alphabet rather than reaching comprehension, and
over first grade three times the odds of stopping at early word reading.
The `from:` terms are origin-status intercepts, several of them for cells
almost nobody occupies, which is what their standard errors say.

The final model was fitted with standard errors, so the loadings now
come with theirs:

```{r loadings-se}
random_intercept_loadings(fit_pov_tr)
```

`random_intercept_scores()` returns each child's estimated position on
the trait, which is the natural quantity to relate to anything else in
the data:

```{r scores, fig.height = 4}
scores <- random_intercept_scores(fit_pov_tr)
boxplot(scores$mean_score ~ ecls_reading$poverty, horizontal = TRUE,
        names = c("Not in poverty", "In poverty"),
        xlab = "Reading readiness (posterior mean factor score)", ylab = "")
```

## Practical notes

* **Sample size.** Muthén & Asparouhov's guidance for binary indicators
  is that RI-LTA does well from about N = 500 with three or more
  occasions, but may need N of 4,000 with only two; for the continuous-
  indicator analogue, Tseng (2024) puts it at upwards of 2,000 cases for
  80% power at a between-profile separation of d = 0.75. The asymmetry
  that makes trying it worthwhile anyway: leaving out a random intercept
  that belongs distorts the statuses and the transitions, while adding one
  that does not belong costs a few parameters that BIC will identify as
  wasted.
* **Starts.** Random-intercept models have a well-known second optimum
  in which the factor takes over the statuses. `fit_lta()`'s restart
  search is built for it, but still read the `Best solution` line, and
  raise `n_init` when it is not replicated.
* **Quadrature.** Check `n_quadrature` the way it is checked above,
  over a range of node counts, and expect the node count to matter more
  once the factor is regressed on covariates.
* **Testing.** Compare random-intercept and ordinary LTA by BIC, never by
  `lr_test()`. Nested covariate models *within* the random-intercept
  family can be tested with `lr_test()` as usual.
* **Not available.** Continuous indicators, random slopes, correlated
  residuals across occasions and second-order (lag-2) transitions are not
  part of this release; the last of these is reported as significant in
  both of the article's own examples, and is a real simplification.

## References

Muthén, B., & Asparouhov, T. (2022). Latent transition analysis with
random intercepts (RI-LTA). *Psychological Methods*, *27*(1), 1–16.
<https://doi.org/10.1037/met0000370>

Kaplan, D. (2008). An overview of Markov chain methods for the study of
stage-sequential developmental processes. *Developmental Psychology*,
*44*(2), 457–467. <https://doi.org/10.1037/0012-1649.44.2.457>

Tseng, M.-C. (2024). Latent profile transition analysis with random
intercepts (RI-LPTA). *Structural Equation Modeling*, *31*(4), 626-634.
<https://doi.org/10.1080/10705511.2023.2284671>
