---
title: "Repeated-Measures LCA: Patterns of Change Over Time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repeated-Measures LCA: Patterns of Change Over Time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

Repeated-measures latent class analysis (RMLCA) treats the *whole
trajectory* of a repeated item as the thing to classify: each latent class
is a typical pattern of responses over time, with no growth-curve shape
imposed (Collins & Lanza, 2010, ch. 7). It is the natural first
longitudinal model when you suspect qualitatively different pathways —
say, "persistent", "declining", and "rare" — rather than variation around
one common curve.

## Simulated example: weekly drinking across four school years

We simulate a cohort of 900 students reporting weekly alcohol use (yes/no)
in each of four school years, generated from three known trajectory
classes so the model has a truth to recover:

* **Rare** (50%): low probability of use at every wave;
* **Escalating** (30%): probability rising steeply across waves;
* **Persistent** (20%): high probability throughout.

```{r simulate}
set.seed(2026)
n   <- 900
cls <- sample(1:3, n, replace = TRUE, prob = c(0.5, 0.3, 0.2))
p   <- rbind(
  c(0.08, 0.10, 0.12, 0.12),   # Rare
  c(0.15, 0.35, 0.60, 0.80),   # Escalating
  c(0.85, 0.85, 0.90, 0.90)    # Persistent
)
drink <- sapply(1:4, function(t) rbinom(n, 1, p[cls, t]))
colnames(drink) <- paste0("drink_y", 1:4)
head(drink)
```

## How many trajectory classes?

`compare_longitudinal()` fits a range of class counts (an RMLCA is fitted
via `fit_rmlca()`, which needs to know how the columns map to occasions —
here four waves of one item):

```{r enumerate, cache = TRUE}
selection <- compare_longitudinal(drink, k_range = 2:4, model = "rmlca",
                                  times = 4, n_init = 10, verbose = FALSE)
```

BIC correctly picks the three classes we generated.

## The three-class solution

```{r fit}
fit <- fit_rmlca(drink, n_classes = 3, times = 4, n_init = 20,
                 random_state = 7)
fit
```

The item-response probabilities per wave — the trajectories themselves —
come from `measurement_summary()`, and `plot()` draws them:

```{r summary}
params <- measurement_summary(fit)
```

```{r plot}
plot(fit, class_labels = c("Rare", "Escalating", "Persistent"),
     main = "Weekly drinking across four school years")
```

The recovered curves match the generating probabilities well: a flat low
class, a steeply rising class, and a flat high class, with sizes close to
the simulated 50/30/20 split.

## Who follows which pathway?

Predictors of trajectory-class membership use the same stepwise workflow
as any other mixtureEM model: hand the fitted model to
`add_covariates()`. We simulate a covariate that genuinely raises the odds
of the Escalating pathway:

```{r covariates}
risk <- rnorm(n, mean = ifelse(cls == 2, 0.8, 0))
fit_cov <- add_covariates(fit, risk)
results <- summary(fit_cov)
```

The escalating class shows the elevated odds ratio for `risk`, as
simulated; the persistent class does not.

## RMLCA or LTA?

RMLCA classifies whole trajectories; latent *transition* analysis (LTA)
instead models wave-to-wave movement between statuses. If your question
is "what pathways exist?", RMLCA answers it directly; if it is "who moves,
when, and toward what?", see the latent transition analysis vignette
(`vignette("lta", package = "mixtureEM")`).

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent
transition analysis: With applications in the social, behavioral, and
health sciences*. Wiley.

Lanza, S. T., & Collins, L. M. (2006). A mixture model of discontinuous
development in heavy drinking from ages 18 to 30: The role of college
enrollment. *Journal of Studies on Alcohol*, *67*(4), 552–561.
<https://doi.org/10.15288/jsa.2006.67.552>
