---
title: "LCA with Complex Survey Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LCA with Complex Survey Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

Large public-health and education surveys are almost never simple random
samples: cases carry *sampling weights* (they stand for different numbers
of people), and they arrive in *strata* and *clusters* (schools, PSUs)
whose members resemble each other. Ignoring the weights biases the
estimates; ignoring the design makes the standard errors too small.
mixtureEM handles both through the `weights`, `strata`, and `cluster`
arguments, available in every model.

## The data: 12 health-risk behaviors, YRBS 2005

The bundled `yrbs2005` dataset holds twelve dichotomous risk-behavior
items for 13,917 U.S. high-school students from the CDC's 2005 Youth Risk
Behavior Survey, together with the survey's weight, PSU, and stratum
variables (`?yrbs2005`). Collins and Lanza (2010, ch. 2 and 4) built
their textbook LCA illustration on these items; we follow their five-class
solution.

```{r data}
data(yrbs2005)
items <- yrbs2005[, 3:14]
names(items)
```

The items contain missing responses; mixtureEM switches affected
indicators to a full-information (FIML) estimator automatically, and a
handful of cases missing on *every* item are dropped with a warning.

## A design-based five-class LCA

```{r fit, cache = TRUE}
set.seed(1)
fit <- suppressWarnings(
  fit_mixture(items, n_classes = 5, measurement = "binary",
              weights = yrbs2005$weight,
              strata  = yrbs2005$stratum,
              cluster = yrbs2005$psu,
              n_init = 20, max_iter = 2000)
)
fit
class_sizes(fit)
```

The five classes reproduce the familiar YRBS typology — a large low-risk
class, substance-experimentation classes, and a small high-everything
class:

```{r plot, fig.height = 5.5}
plot(fit, main = "Health-risk behavior classes (YRBS 2005)")
```

```{r params}
params <- measurement_summary(fit)
```

## Predictors under the survey design

The stored design travels with the fitted model, so the stepwise
covariate analysis is design-based too: the standard errors below are
linearized (sandwich) estimates that respect the strata and clusters.

```{r covariates, cache = TRUE}
fit_cov <- add_covariates(fit, yrbs2005[, c("grade", "sex")])
results <- summary(fit_cov)
```

Note the `Standard errors:` line naming the estimator: with a cluster
sample, a model that ignored the design would report noticeably narrower
intervals — narrower than the data warrant.

## What the weights do and do not change

Two practical points that trip up survey LCA:

* **Weight scale.** Sampling weights are rescaled internally so that they
  sum to the number of observed cases; only their *relative* sizes
  matter, and the effective sample size in BIC stays the number of cases
  actually collected. Frequency weights (`weight_type = "frequency"`,
  where one row stands for that many identical cases) are used as-is.
* **Class enumeration.** Information criteria remain comparable across
  class counts under a fixed design, so `compare_mixtures()` works
  unchanged — pass the same `weights`/`strata`/`cluster` to every
  candidate model.

## References

Collins, L. M., & Lanza, S. T. (2010). *Latent class and latent
transition analysis: With applications in the social, behavioral, and
health sciences*. Wiley.
