---
title: "Infidelity Behavior Patterns: LCA with Covariates and a Distal Outcome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Infidelity Behavior Patterns: LCA with Covariates and a Distal Outcome}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(mixtureEM)
```

## Background

This vignette walks through Ventura-León et al. (2025), *"Exploring Infidelity
Behavior Patterns in a Sample of Peruvian Young Adults: A Latent Class
Analysis,"* using the `ventura_leon` dataset bundled with mixtureEM
(`?ventura_leon`).

Four hundred Peruvian young adults (75.3% women, mean age 25.3) answered 16
binary items from the Multidimensional Infidelity Inventory-S (MII-S), each
asking whether they had engaged in a particular unfaithful thought or
behavior. The original study used latent class analysis (LCA) to identify
subgroups with distinct patterns, then related class membership to sex, age,
sexual orientation, and relationship duration.

We reproduce the class solution first, then improve on the original
covariate analysis: the paper assigned each respondent to their single most
likely class and ran chi-square/ANOVA tests on those labels, which ignores
classification uncertainty. mixtureEM's `add_covariates()` instead uses a
bias-corrected multinomial logistic regression (Vermunt, 2010) that
properly accounts for it.

Data are shared here with permission from the author team (one of whom,
Pablo D. Valencia, also maintains mixtureEM); see `?ventura_leon` for the
full source and license note.

## The data

```{r data}
data(ventura_leon)
items <- ventura_leon[, 7:22]   # the 16 infidelity items
names(items)
```

The 16 items are already binary (0 = never endorsed, 1 = endorsed), with
short informative names; see `?ventura_leon` for each item's exact wording.

## How many classes?

The paper compared 1- through 10-class solutions by BIC (their Table 2) and
selected 4 classes. We do the same here.

```{r model-selection, cache = TRUE}
selection <- compare_mixtures(X = items, measurement = "binary",
                               k_range = 1:6, n_init = 20)
```

As in the paper, BIC keeps improving for a few classes past 4 (a common
pattern in LCA, where the BIC curve is often quite flat near its minimum);
we follow the paper in choosing 4 classes for interpretability, since it
recovers a very clean, well-separated typology (see below).

## Fitting the 4-class model

```{r fit4}
set.seed(1)
fit <- fit_mixture(items, n_classes = 4, measurement = "binary",
                    n_init = 30, max_iter = 2000)
fit
```

### Comparing to the published solution

The paper's four classes were: *Fidelity* (42.5%, low on everything),
*Affective interest* (27.2%, moderate on thought/attraction items),
*Infidelity* (15.7%, high on nearly every item), and *Sexual desire*
(14.4%, high on desire/attraction items but not overt behavior). Classes
are sorted by size, so `class_sizes()` lines up with that ordering:

```{r compare-prevalence}
class_names <- c("Fidelity", "Affective interest", "Infidelity", "Sexual desire")
class_sizes(fit)
```

The item-response probabilities per class come from
`measurement_summary()`, which prints the full table and also returns it as
a data frame for further use:

```{r compare-profile}
params <- measurement_summary(fit)
```

Because the returned table is an ordinary data frame, follow-up questions
take one line — for example, which behaviors the third-largest class
endorses with high probability:

```{r profile-followup}
subset(params, class == 3 & estimate > 0.5)
```

```{r plot4}
plot(fit, class_labels = class_names,
     main = "Infidelity behavior patterns (4-class LCA)")
```

## Covariates: a stronger analysis than the original paper

The paper related class membership to sex, age, sexual orientation, and
relationship duration by assigning each respondent to their single most
likely class (the modal posterior) and then testing associations with
separate chi-square tests and an ANOVA — a "classify-and-analyze" approach
that treats class membership as if it were observed without error, which
biases the association estimates toward the null (Bakk et al., 2014).

Instead, we hand the chosen model to `add_covariates()`, which runs the
bias-adjusted three-step analysis (Vermunt, 2010) on the *same* solution we
just inspected — the measurement model is not re-estimated, so the classes
cannot shift under our feet:

```{r covariates}
covariates <- ventura_leon[, c("sex", "age", "sexual_orientation",
                                "relationship_duration")]
fit_cov <- add_covariates(fit, covariates)
results <- summary(fit_cov)
```

The `Standard errors:` line records which variance estimator produced the
intervals. The default also carries the uncertainty in the step-1 class
solution into the step-3 coefficients (Bakk et al., 2014):
treating the classes as if they had been observed rather than estimated
makes the intervals too narrow. See `?covariate_se`.

`summary()` also returns its tables invisibly, so the odds ratios are
available as a data frame:

```{r covariate-table}
head(results$coefficients)
```

Reference class 1 is Fidelity (the largest class). Reading the odds ratios
against that reference:

* **Sex.** Women have higher odds of Affective interest relative to
  Fidelity, and much lower odds of Infidelity relative to Fidelity — in
  line with the paper's finding that "men were more likely to belong to the
  sexual desire class and women to the emotional interest and fidelity
  classes."
* **Age.** Older respondents have higher odds of Infidelity relative to
  Fidelity, matching the paper's finding that age was significantly
  related to class membership.
* **Sexual orientation.** One row looks like a discovery the paper's
  classify-and-analyze approach missed: non-heterosexual respondents have
  higher odds of Sexual desire relative to Fidelity (OR = 2.75,
  p = .022). But the omnibus test for sexual orientation — the one built
  to ask whether the covariate distinguishes *any* pair of classes at all
  — is not significant here either (Wald chi-square = 6.44, df = 3,
  p = .092), matching the paper's own null result. With three pairwise
  contrasts tested, a single row below .05 is not strong evidence on its
  own; read the omnibus row first, and treat an isolated significant
  contrast beside a non-significant omnibus test as a hypothesis worth a
  better-powered look, not a confirmed finding.
* **Relationship duration.** Neither approach finds a significant
  association, consistent with the paper.

This is a good illustration of why mixtureEM applies a bias correction
whenever classes are related to external variables: the "obvious" approach
of classifying and then testing is only valid when classification is
(almost) perfect, which is rarely true in practice.

## A distal outcome: mean age by class

Covariates ask "who ends up in which class?". The complementary question —
"how do the classes differ on some outcome?" — is a *distal outcome*
analysis, and it reuses the same fitted model through `add_outcome()`.
Here we describe the classes by their mean age (the BCH correction is the
default for a continuous outcome; Bakk & Vermunt, 2016). Age already
entered the model as a covariate above, so treating it as a distal outcome
too does not answer a substantively new question — we reuse it purely
because it is a continuous variable already at hand, to keep the example
self-contained; a real distal-outcome analysis would pick a variable the
classes are actually expected to predict.

```{r distal}
fit_age <- add_outcome(fit, ventura_leon$age)
age_results <- summary(fit_age)
age_results$outcome$means
```

The omnibus Wald test in the printed output asks whether the class means
differ at all, before the per-class table is read.

## References

Bakk, Z., Oberski, D. L., & Vermunt, J. K. (2014). Relating latent class
assignments to external variables: Standard errors for correct inference.
*Political Analysis*, *22*(4), 520–540. <https://doi.org/10.1093/pan/mpu003>

Bakk, Z., & Vermunt, J. K. (2016). Robustness of stepwise latent class
modeling with continuous distal outcomes. *Structural Equation Modeling*,
*23*(1), 20–31. <https://doi.org/10.1080/10705511.2014.955104>

Ventura-León, J., Reyes, A., Valencia, P. D., Tocto-Muñoz, S.,
Gamboa-Melgar, G., Ruiz-Castro, J., & Lino-Cruz, C. (2025). Exploring
infidelity behavior patterns in a sample of Peruvian young adults: A latent
class analysis. *Journal of Marital and Family Therapy*, *51*, e70066.
<https://doi.org/10.1111/jmft.70066>

Vermunt, J. K. (2010). Latent class modeling with covariates: Two improved
three-step approaches. *Political Analysis*, *18*(4), 450–469.
<https://doi.org/10.1093/pan/mpq025>
